---
title: "Applications and Portable Device Execution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Applications and Portable Device Execution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(torch)
library(riemtorch)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = torch::torch_is_installed())
if (torch::torch_is_installed()) {
  torch_manual_seed(12)
  torch_set_num_threads(1)
  device <- riem.device()
  device
}
```

`riemtorch` selects one compatible device for each managed run. It preserves the
requested precision, so an accelerator that cannot execute float64 is skipped
rather than used with a silent conversion. Pass `device = "cpu"` to reproduce a
run on the host, or `device = "cuda:1"` to select the second visible CUDA GPU.
`riem.devices()` reports the devices that the current R process can see and the
result of a small compatibility probe. Resolution follows an explicit argument,
`options(riemtorch.device = ...)`, `RIEMTORCH_DEVICE`, then automatic discovery.
An explicit `device = "auto"` asks for discovery even when a default is set.
Scheduler visibility restrictions are respected because only devices exposed to
the R process are considered.

```{r inventory}
riem.devices()
```

## Principal subspace estimation

The leading two-dimensional eigenspace of a covariance matrix is a point on a
Grassmann manifold. Registering the covariance as problem data lets the execution
driver move it with the initial point.

```{r pca}
X <- scale(as.matrix(iris[, 1:4]), center = TRUE, scale = FALSE)
C <- torch_tensor(crossprod(X) / nrow(X), dtype = torch_float64())
G <- manifold.grassmann(4, 2)
pca_problem <- riem.problem(G,
  function(Q, data) -torch_trace(Q$t()$matmul(data)$matmul(Q)), data = C)
pca_fit <- riem.optimize(pca_problem, riem.random(G), "conjugate_gradient",
  control = list(max_iterations = 15))
pca_fit$objective
pca_fit$constraint_residuals
```

## Low-rank matrix completion

The mask and observations are ordinary registered tensors. Changing the loss
does not require changing the fixed-rank geometry or solver.

```{r completion}
truth <- torch_tensor(outer(1:5, c(1, -1, 2)) + outer(c(1, 0, -1, 2, 1),
  c(2, 1, -1)), dtype = torch_float64())
mask <- torch_tensor(matrix(c(1,1,0, 1,0,1, 0,1,1, 1,1,0, 1,0,1), 5, 3,
  byrow = TRUE), dtype = torch_float64())
F <- manifold.fixedrank(5, 3, 2)
completion <- riem.problem(F, function(Z, data) {
  observed <- data$mask * (Z - data$truth)
  torch_sum(observed^2) / 2
}, data = list(mask = mask, truth = truth))
completion_fit <- riem.optimize(completion, riem.random(F), "conjugate_gradient",
  control = list(max_iterations = 10))
completion_fit$constraint_residuals
```

## An affine-invariant SPD mean

The Fréchet mean minimizes a sum of squared affine-invariant distances. This
example uses only the general problem interface and SPD geometry.

```{r spd-mean}
S <- manifold.spd(2, "airm")
A <- torch_diag(torch_tensor(c(1, 4), dtype = torch_float64()))
B <- torch_diag(torch_tensor(c(9, 1), dtype = torch_float64()))
mean_problem <- riem.problem(S, function(X, data) {
  (riem.sqdist(S, X, data$A) + riem.sqdist(S, X, data$B)) / 2
}, data = list(A = A, B = B))
mean_fit <- riem.optimize(mean_problem, riem.random(S), "conjugate_gradient",
  control = list(max_iterations = 12))
mean_fit$point
```

## Orthogonality-constrained torch training

`riem.train()` moves the module before constructing the optimizer, transfers
each minibatch, and retains the actual parameter objects used by the optimizer.
Here the learned linear map remains on a Stiefel manifold after every update.

```{r orthogonal-training}
inputs <- torch_randn(12, 4, dtype = torch_float64())
teacher <- torch_eye(4, dtype = torch_float64())[, 1:2]
responses <- inputs$matmul(teacher)
loader <- dataloader(tensor_dataset(inputs, responses), batch_size = 4)
frame_geometry <- manifold.stiefel(4, 2, "euclidean")
orthogonal_module <- nn_module("orthogonal_module",
  initialize = function() {
    self$frame <- nn_parameter(riem.random(frame_geometry, device = "cpu"))
  },
  forward = function(x) x$matmul(self$frame))
trained <- riem.train(orthogonal_module(), loader,
  loss = function(model, batch)
    (model(batch[[1]]) - batch[[2]])$square()$mean(),
  optimizer_factory = function(model)
    optim_rsgd(list(model$frame), lr = 0.05, manifold = frame_geometry),
  epochs = 2)
riem.belongs(frame_geometry, trained$model$frame)
trained$execution
```

The selected device and dtype are stored in every managed result. For a forced
CPU run on a GPU workstation, specify the override on the run itself:

```{r cpu-override}
cpu_fit <- riem.optimize(pca_problem, riem.random(G, device = "cpu"),
  "steepest_descent", control = list(max_iterations = 2), device = "cpu")
cpu_fit$execution
```
