---
title: "Checking geometry and derivatives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking geometry and derivatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", eval=torch::torch_is_installed())
library(riemtorch)
library(torch)
if(torch_is_installed()) {torch_set_num_threads(1); torch_manual_seed(13)}
```

A diagnostic checks a mathematical identity at sampled points. Repeat it at
several interior points, and inspect the errors rather than treating a passing
result as a proof over the whole manifold.

```{r}
M <- manifold.sphere(3)
x <- torch_tensor(c(0.6, 0.8, 0), dtype=torch_float64())
u <- torch_tensor(c(-0.8, 0.6, 0), dtype=torch_float64())
P <- riem.problem(M, function(x) -x[1])
riem.check.manifold(M,x,u)$checks
riem.check.gradient(P,x,u)$curves
riem.check.hessian(P,x,u)$checks
```

The gradient Taylor remainder should decay quadratically until roundoff
intervenes. A Hessian Taylor test requires an exponential map or a declared
second-order retraction. Without one, the Taylor test is inconclusive while
linearity and symmetry tests still run.

```{r}
bad <- riem.problem(M,function(x) -x[1],rgrad=function(x) x*0)
stopifnot(riem.check.gradient(bad,x,u)$status == "fail")
Q <- riem.problem(manifold.euclidean(3),function(x) (x*x)$sum())
riem.hessian.spectrum(Q,torch_ones(3,dtype=torch_float64()))$values
riem.devices(operations="basic")
```

Spectrum results are Ritz estimates with residuals, not certificates that no
unseen more extreme eigenvalue exists. Device probes qualify specific operations;
unknown user callbacks retain the conservative full probe.
