---
title: "Optimization with complex geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimization with complex geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>",eval=torch::torch_is_installed())
library(riemtorch);library(torch)
if(torch_is_installed()) {torch_set_num_threads(1);torch_manual_seed(43)}
```

Complex points use a real Hermitian metric, Re(trace(U* V)). Objectives must
return real scalars: float64 for complex128 points, float32 for complex64.
Torch's conjugate gradient convention is checked using independent real and
imaginary perturbations. Complex Euclidean, sphere and phase geometries support
automatic Hessians; complex frame and unitary geometries currently qualify only
first-order optimization. Matrix factorizations at repeated singular values
are not advertised as automatically twice differentiable.

```{r}
M <- manifold.complexcircle(4)
x <- riem.random(M,device="cpu")
P <- riem.problem(M,function(z) -z$real$sum())
riem.check.gradient(P,x)$checks
riem.check.hessian(P,x)$checks
fit <- riem.optimize(P,x,device="cpu")
fit$objective
riem.belongs(M,fit$point)
```

Complex Stiefel uses the embedded Euclidean metric. Complex Grassmann uses
orthonormal frames and objectives must be invariant to unitary basis changes.
Its logarithm is local and rejects the cut locus.

```{r}
G <- manifold.grassmann(4,2,field="complex")
X <- riem.random(G,device="cpu")
U <- riem.tangent(G,X,torch_randn_like(X))*0.05
Y <- riem.exp(G,X,U)
(riem.log(G,X,Y)-U)$abs()$max()
riem.belongs(manifold.unitary(2),riem.random(manifold.unitary(2),device="cpu"))
```
