---
title: "Constraints and intrinsic nonsmooth optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constraints and intrinsic nonsmooth optimization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE,comment="#>",eval=torch::torch_is_installed())
library(riemtorch);library(torch)
if(torch_is_installed()) torch_set_num_threads(1)
```

Equality constraints use h(x)=0; inequalities use g(x)<=0. The augmented
Lagrangian reuses smooth inner solvers, tightens their requested accuracy and
updates multipliers. Its result reports primal feasibility, Lagrangian gradient
norm, complementarity, dual feasibility, and every inner-solver status.

```{r}
P <- riem.problem.constrained(manifold.euclidean(2),
  function(x) ((x-2)^2)$sum()/2,
  equality=function(x) x$sum()$reshape(1)-1,
  inequality=function(x) -x)
fit <- riem.optimize(P,torch_tensor(c(0.5,0.5),dtype=torch_float64()),
  "augmented_lagrangian",device="cpu")
fit$point
fit$multipliers
fit$kkt
```

Automatic constraint adjoints use torch differentiation. Explicit metric
adjoints may be supplied as `equality_adjoint(x, weights)` and
`inequality_adjoint(x, weights)`. Registered data is passed consistently to all
callbacks. Inner failures, work budgets and user stops do not count as KKT
convergence. Even successful KKT tests are local conditions, not global optimality.

For a smooth function f and nonsmooth h, the intrinsic proximal callback must
solve

\[
\mathrm{prox}_{t h}(q)=\arg\min_{y\in M}\left(h(y)+\frac{d(y,q)^2}{2t}\right).
\]

Projection after ambient shrinkage generally solves a different problem. The
unaccelerated method first takes an exponential gradient step, applies that
proximal map, then checks sufficient decrease. The norm of the intrinsic
gradient mapping is its residual; a small ambient gradient is not used.

```{r}
soft <- function(x,t) x$sign()*(x$abs()-t)$clamp(min=0)
P <- riem.problem.composite(manifold.positive(2),
  function(x) ((x$log()-2)^2)$sum()/2,
  nonsmooth=function(x) x$log()$abs()$sum(),
  prox=function(q,step) soft(q$log(),step)$exp())
fit <- riem.optimize(P,torch_ones(2,dtype=torch_float64()),device="cpu")
fit$point$log() # solution is c(1,1) in log coordinates
fit$proximal_residual
```

Euclidean, affine, log-Euclidean positive/SPD, and hyperbolic geometries and their
products/scalings are initially qualified. Qualification concerns the geometry
contract, not arbitrary user proximal callbacks. Other geometries require
`qualified=FALSE` and remain experimental. The caller must validate its prox
against optimality conditions for its particular h.

Cyclic proximal point takes lists of value functions and matching proximal
maps, without a separate smooth term. Default steps decay with exponent 0.75.
Its cycle displacement divided by the step is reported, but a small cycle
residual alone is not a stationarity certificate for the sum.

```{r}
C <- riem.problem.composite(manifold.euclidean(1),
  nonsmooth=list(function(x) x$square()$sum()/2,
                 function(x) ((x-2)^2)$sum()/2),
  prox=list(function(q,t) q/(1+t),function(q,t) (q+2*t)/(1+t)))
fit <- riem.optimize(C,torch_tensor(3,dtype=torch_float64()),
  "cyclic_proximal_point",control=list(max_iterations=50),device="cpu")
fit$point
fit$termination
```
