---
title: "Custom tensor objectives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom tensor objectives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

The solver only sees a scalar torch objective and a manifold. A new loss does
not require a new estimator. Here the exact minimizer is a normalized vector.

```{r}
M <- manifold.sphere(3)
a <- torch_tensor(c(1, 2, 3), dtype = torch_float64())
x <- torch_tensor(c(1, 0, 0), dtype = torch_float64())
P <- riem.problem(M, function(x, data) -torch_sum(x * data), data = a)
fit <- riem.optimize(P, x, "conjugate_gradient")
fit
as.numeric(fit$point)
stopifnot(abs(fit$objective + sqrt(14)) < 1e-7)
```

An ambient gradient and an intrinsic gradient are different callbacks. The
sphere gradient removes the radial component; the package performs that
conversion for an ambient callback.

```{r}
analytic <- riem.problem(M, function(x, data) -torch_sum(x * data),
                        egrad = function(x, data) -data, data = a)
riem.evaluate(analytic, x)$gradient
```

Objectives must return one scalar tensor with the point's dtype/device. Data can
be converted explicitly at the boundary, then all numerical work stays in torch.
Deterministic callbacks must not change batches, dropout masks or other state
inside a line search. Returned solutions are detached from the optimization graph.
