---
title: "Deterministic optimization and diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deterministic optimization and diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

Orthogonal Procrustes is a nontrivial matrix objective with a known solution.
Neither the geometry nor the solver contains a special Procrustes estimator.

```{r}
M <- manifold.stiefel(5, 2, "euclidean")
A <- torch_randn(c(5, 2), dtype = torch_float64())
P <- riem.problem(M, function(X, data) -torch_sum(data * X), data = A)
fit <- riem.optimize(P, riem.random(M), "conjugate_gradient",
                     control = list(max_iterations = 150))
fit
fit$evaluations
stopifnot(riem.belongs(M, fit$point))
reference <- riem.project(M, riem.to(A, device = fit$execution$device,
                                     dtype = fit$execution$dtype))
torch_sum((fit$point - reference)^2)$item()
```

Armijo backtracking rejects infeasible or nonfinite trials. Strong Wolfe is
available only where the true retraction-curve derivative is supported. PR+
conjugate gradient restarts on loss of descent. L-BFGS transports secant history
and skips invalid curvature pairs; BB bounds its spectral step lengths.

`converged_gradient` is distinct from `stopped_small_step`, `max_iterations`,
`line_search_failed`, `domain_error`, `nonfinite_objective`, and
`numerical_failure`. Always inspect termination and the available gradient norm.
A feasible iterate is useful even when the maximum iteration budget was reached.

```{r}
head(riem.support("solvers"))
fit$diagnostics$restarts
```

The four derivative-free searches and cubic regularization are implemented but
marked experimental. They have explicit algorithm variants in the installed
mathematical notes. Heuristics do not claim gradient convergence.
