---
title: "Adding a geometry without changing a solver"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a geometry without changing a solver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

A custom geometry supplies point-level tensor kernels through `riem.manifold()`.
This positive line is Euclidean in log coordinates and has metric u*v/x^2.

```{r}
M <- riem.manifold("positive_line", shape = 1, dimension = 1,
  metric = "log_pullback", operations = list(
    belongs = function(x, tol) x$item() > 0,
    tangent = function(x, u) u,
    inner = function(x, u, v) torch_sum(u * v / x^2),
    egrad2rgrad = function(x, u) x^2 * u,
    retr = function(x, u) x * torch_exp(u/x),
    transport = function(x, u, y, v) v * y/x,
    project = function(x) torch_abs(x) + 0.01
  ), capabilities = list(isometric_transport = TRUE))
P <- riem.problem(M, function(x) torch_sum((torch_log(x) - log(3))^2)/2)
x <- torch_tensor(1, dtype = torch_float64())
fit <- riem.optimize(P, x, "conjugate_gradient")
fit
stopifnot(abs(fit$point$item() - 3) < 1e-8)
```

Membership and tangency, metric duality, R_x(0)=x, and D R_x(0)[u]=u are the
minimum contract. Test directional derivatives at several steps, noncontiguous
inputs, dtype/device preservation, and a nontrivial objective. Transport must
own vectors at the right base point and declare its path. Exact geodesic and
higher-order capabilities are optional; do not declare them merely because a
forward formula exists. Geometry kernels cannot call another numerical backend.
