---
title: "Geometry, metrics and representations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geometry, metrics and representations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

Constructors specify a space, metric and representation. Multi-metric families
require an explicit choice. SPD tangents are symmetric matrices; choosing AIRM
changes gradient conversion, not just the reported distance.

```{r}
X <- torch_diag(torch_tensor(c(2, 3), dtype = torch_float64()))
G <- torch_eye(2, dtype = X$dtype)
M <- manifold.spd(2, "airm")
riem.egrad2rgrad(M, X, G)       # X G X
riem.inner(M, X, G, G)
riem.capabilities(M)
```

Leading batch axes are independent geometry evaluations. They are never inferred
to be factors of a joint optimization problem. Every input must have identical
batch axes; implicit broadcasting is deliberately rejected.

```{r}
S <- manifold.sphere(3)
batch <- riem.random(S, batch_shape = c(2, 4))
u <- riem.tangent(S, batch, torch_randn_like(batch))
riem.inner(S, batch, u, u)$shape
stopifnot(all(riem.belongs(S, riem.retr(S, batch, u, step = 0.1))))
```

Grassmann frame objectives must be invariant under changes of orthonormal basis.
PSD factor objectives must depend on Y Y'. Kendall shape objectives must be
invariant under rotations. Low-rank and shape constructors document their regular
strata. Repair through `riem.project()` is explicit; a solver never repairs a
failed trial invisibly. The installed geometry ledger maps all source rows:

```{r}
head(riem.support("geometry"), 8)
```
