---
title: "Numerical domains and reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Numerical domains and reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

CPU float64 is the reference validation platform. A managed run probes visible
devices and preserves its requested precision: compatible CUDA devices are
considered first, then compatible MPS, then CPU. This is runtime qualification,
not evidence that an accelerator is faster or numerically identical. CUDA and
MPS were unavailable in the local reference environment. Use `device = "cpu"`
for a reproducible host override; an unavailable explicit accelerator is an
error rather than a request to fall back silently.

Batch geometry is supported with exact shape matching; independent batched
solves are not implicitly reduced to one loss. Native kernels cover the most
frequently used vector, frame, Grassmann and SPD geometries. Explicit host scalar
extraction is still needed for stopping and validation and may synchronize an
accelerator. An out-of-memory or numerical failure after execution starts does
not restart the run on another device.

Smooth matrix functions should be differentiated as matrices, not through
arbitrary eigenvector gauges. The matrix-log custom backward is finite at an
identity spectrum.

```{r}
X <- torch_eye(2, dtype = torch_float64(), requires_grad = TRUE)
value <- riem.matrix.function(X, "log")$sum()
value$backward()
X$grad
```

These custom log/root backward rules are first-order only. An exact HVP through
them requires an analytic callback. Do not call external double backward on
these primitives and assume it implements the full matrix second derivative.
Rank projection and quotient charts have additional domains in their help pages.

R torch 0.17.0 can promote a numeric scalar on the **left** of a zero-dimensional
tensor operation through the default float dtype. For strict float64 objectives,
write `tensor * 0.1` instead of `0.1 * tensor`, or construct a matching torch
constant. riemtorch's scalar metric arithmetic follows this convention.

An example based on R's built-in iris data demonstrates the registered-data
boundary and an SPD objective. The data is converted once; the execution driver
moves the registered target with the point. Tensors captured privately in a
callback closure are not discoverable and remain the caller's responsibility.

```{r}
X <- torch_tensor(as.matrix(iris[, 1:4]), dtype = torch_float64())
X <- X - X$mean(dim = 1, keepdim = TRUE)
C <- X$t()$matmul(X)/(nrow(iris) - 1) + torch_eye(4, dtype = X$dtype)*0.1
T <- riem.matrix.function(C, "log")
M <- manifold.spd(4, "lerm")
P <- riem.problem(M, function(x, data) {
  torch_sum((riem.matrix.function(x, "log") - data)^2) / 2
}, data = T)
fit <- riem.optimize(P, torch_eye(4, dtype = X$dtype), device = "cpu")
fit
stopifnot(fit$objective < 1e-10)
```

Finite-sum SGD performs full objective and gradient evaluations initially and
finally. Intermediate iterations use minibatches unless a positive
`full_evaluation_every` requests periodic full diagnostics. `batch_fn` avoids
per-term R calls, and chunked full evaluations bound the size of each computation
graph; they do not remove the total cost of a requested full diagnostic.

Version-two checkpoints can include model parameters and buffers, caller-owned
progress or data-order metadata, and R, torch CPU and visible CUDA RNG states.
Exact continuation still depends on matching objective code, runtime behavior
and minibatch order. Results resumed across CPU and CUDA need not be bitwise
identical. Custom geometry callbacks are reconstructed by the program and are
never deserialized from a checkpoint.
