---
title: "Hessians and least-squares problems"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hessians and least-squares problems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

Exact Hessian-vector products include the metric connection correction. A sphere
quadratic illustrates why projecting an ambient Hessian is insufficient: a
constant-on-sphere objective has zero intrinsic Hessian.

```{r}
M <- manifold.sphere(3)
P <- riem.problem(M, function(x) torch_sum(x^2)/2)
x <- torch_tensor(c(1, 0, 0), dtype = torch_float64())
u <- torch_tensor(c(0, 1, 0), dtype = torch_float64())
riem.hessian(P, x, u)
```

Trust regions and Newton-CG use matrix-free exact HVPs. For unsupported automatic
geometry/objective combinations, provide `rhess(x,u)` explicitly. Custom matrix
logarithm/root operations support first derivatives only and are rejected in
automatic higher-order paths.

```{r}
E <- manifold.euclidean(2)
P <- riem.problem(E, function(x) torch_sum((x - 1)^2)/2)
fit <- riem.optimize(P, torch_zeros(2, dtype = torch_float64()), "trust_regions")
fit
fit$diagnostics$subproblems
```

Least-squares problems use the selected metric adjoint J*, not an unqualified
coordinate transpose. The residual weights are fixed positive semidefinite.
Gauss--Newton and LM advertise the approximation J* W J; LM adds lambda I in
the tangent metric and adapts damping by actual/predicted reduction.

```{r}
P <- riem.problem.leastsquares(E, function(x) x - 1,
                              weights = torch_tensor(c(1, 3), dtype = torch_float64()))
fit <- riem.optimize(P, torch_zeros(2, dtype = torch_float64()), "levenberg_marquardt")
fit
stopifnot(fit$objective < 1e-10)
```
