---
title: "Products, blocks and finite sums"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Products, blocks and finite sums}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = torch::torch_is_installed())
library(riemtorch)
library(torch)
if (torch::torch_is_installed()) {
  torch::torch_set_num_threads(1)
  torch::torch_manual_seed(1)
}
```

Named products express joint parameters. Fixed metric weights affect gradient
conversion by reciprocal scaling, and are not loss weights.

```{r}
M <- manifold.product(list(position = manifold.euclidean(2),
                          direction = manifold.sphere(3)), weights = c(2, 5))
x <- list(position = torch_tensor(c(2, -1), dtype = torch_float64()),
          direction = torch_tensor(c(1, 0, 0), dtype = torch_float64()))
P <- riem.problem(M, function(z) torch_sum(z$position^2)/2 - z$direction[2])
fit <- riem.optimize(P, x, "alternating_gradient",
                     control = list(max_iterations = 100))
fit
stopifnot(riem.belongs(M, fit$point))
```

A finite-sum term callback identifies each observation by an index. The loss
normalization is explicit, including minibatch rescaling for full sums.

```{r}
observations <- torch_tensor(1:5, dtype = torch_float64())
P <- riem.problem.finitesum(manifold.euclidean(1),
  function(x, i, data) torch_sum((x - data[i])^2)/2,
  batch_fn = function(x, indices, data) {
    (x - data[indices])$square()$reshape(c(length(indices))) / 2
  },
  data = observations, n = 5, normalization = "mean",
  evaluation_batch_size = 2)
x <- torch_zeros(1, dtype = torch_float64())
fit <- riem.optimize(P, x, "stochastic_gradient", control = list(
  max_iterations = 30, batch_size = 5, step_size = 0.2))
fit
fit$diagnostics$sampling
```

The stochastic solver evaluates minibatches after each update. Full objective and
gradient evaluations are chunked and occur initially and finally. Set
`full_evaluation_every` to a positive interval for periodic stationarity checks;
minibatch gradient norms alone are never reported as convergence certificates.
For a caller-controlled dataloader, use `riem.train()` with a native training
optimizer.
