---
title: "Average covariance matrices geometrically"
description: "Compare log-Euclidean and affine-invariant averages of covariance matrices from European stock-index returns."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(riemtorch)
library(torch)
if (!torch_is_installed()) {
  stop("This worked example requires the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(2026)
set.seed(2026)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4, fig.align = "center", dpi = 120)
```

Covariance matrices describe both variation and relationships among variables.
Their geometric average depends on the distance used. This example forms
covariance matrices from successive blocks of historical stock-index returns,
then minimizes squared distances under two SPD metrics.

For a smaller introduction, see [the two-matrix SPD example](example-spd-mean.html).
The [geometry and metrics guide](geometry-and-metrics.html) introduces
`manifold.spd()`, `riem.sqdist()`, and `riem.log()`. The application illustrates
how the choice of geometry changes a covariance summary.

## Data and covariance blocks

R's [`EuStockMarkets` dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/EuStockMarkets.html)
records 1,860 successive business-time observations for DAX, SMI, CAC, and FTSE
from 1991--1998. We use percentage log returns, $100\log(P_t/P_{t-1})$, rather
than prices. The data have no missing values. See
[Data sources and reproducibility](example-data-sources.html) for attribution.

Divide the 1,859 return observations into 15 nonoverlapping blocks of 120;
discard the final 59 returns so that all blocks have equal size. For each sample
covariance $C$, use $0.95C+0.05\operatorname{tr}(C)I/4$. This fixed shrinkage
stabilizes eigenvalues while preserving trace. It is a preprocessing choice,
not a tuning parameter selected from these results.

```{r data}
prices <- as.matrix(datasets::EuStockMarkets)
returns <- apply(log(prices), 2, diff) * 100
block_size <- 120L
block_count <- nrow(returns) %/% block_size
blocks <- lapply(seq_len(block_count), function(i) {
  rows <- (i - 1L) * block_size + seq_len(block_size)
  covariance <- cov(returns[rows, , drop = FALSE])
  covariance * 0.95 + diag(sum(diag(covariance)) / ncol(returns),
                           ncol(returns)) * 0.05
})
stopifnot(!anyNA(returns), block_count == 15,
          all(vapply(blocks, function(C) min(eigen(C, symmetric = TRUE)$values) > 0,
                     logical(1))))
data.frame(return_observations = nrow(returns), complete_blocks = block_count,
           observations_per_block = block_size,
           discarded_returns = nrow(returns) %% block_size)
```

## Objective and metric

For either distance $d$, solve

\[
 \min_{G\in\mathrm{SPD}(4)} \frac{1}{2B}\sum_{b=1}^{B}d(G,C_b)^2.
\]

The log-Euclidean metric measures the Frobenius distance between matrix
logarithms. The affine-invariant metric measures the logarithm after whitening
one matrix by the other. Both keep iterates positive definite, but they give
different means. For either metric the Riemannian gradient of this objective is
$-B^{-1}\sum_b \operatorname{Log}_G(C_b)$, allowing an explicit gradient callback.

```{r problems}
tensor_blocks <- lapply(blocks, torch_tensor, dtype = torch_float64(), device = "cpu")
make_problem <- function(metric) {
  M <- manifold.spd(4, metric = metric)
  riem.problem(
    M,
    function(G, data) {
      Reduce(`+`, lapply(data, function(C) riem.sqdist(M, G, C))) / (2 * length(data))
    },
    rgrad = function(G, data) {
      -Reduce(`+`, lapply(data, function(C) riem.log(M, G, C))) / length(data)
    },
    data = tensor_blocks
  )
}
problems <- list(log_euclidean = make_problem("lerm"),
                 affine_invariant = make_problem("airm"))
arithmetic <- Reduce(`+`, blocks) / length(blocks)
initial <- torch_tensor(arithmetic, dtype = torch_float64(), device = "cpu")
```

All data tensors are registered with their problems, so moving a solve to a
different device also prepares these matrices on that device. The arithmetic
mean provides the same valid starting point for both optimizations.

```{r solve}
fits <- lapply(problems, function(problem) {
  riem.optimize(problem, initial, method = "steepest_descent", device = "cpu",
                control = list(max_iterations = 60, gradient_tolerance = 1e-7))
})
means <- lapply(fits, function(fit) as.matrix(fit$point))
data.frame(metric = names(fits),
           objective = vapply(fits, function(fit) fit$objective, numeric(1)),
           iterations = vapply(fits, function(fit) fit$iterations, numeric(1)),
           gradient_norm = vapply(fits, function(fit) fit$gradient_norm, numeric(1)),
           termination = vapply(fits, function(fit) fit$termination, character(1)))
```

Each objective uses its own metric, so a smaller value across these two rows
does not by itself indicate a better fit. Check the termination reason as well
as the gradient norm: an iteration budget is not mathematical convergence.

## Independent numerical checks

For the log-Euclidean metric, the exact solution is
$\exp\{B^{-1}\sum_b\log C_b\}$. Recalculate it with base R's symmetric
eigendecomposition, independently of torch's matrix operations.

```{r log-reference}
matrix_function <- function(C, fun) {
  decomposition <- eigen((C + t(C)) / 2, symmetric = TRUE)
  decomposition$vectors %*% diag(fun(decomposition$values)) %*%
    t(decomposition$vectors)
}
log_average <- Reduce(`+`, lapply(blocks, matrix_function, fun = log)) / length(blocks)
log_reference <- matrix_function(log_average, exp)
log_reference_error <- norm(means$log_euclidean - log_reference, "F") /
  norm(log_reference, "F")
```

For the affine-invariant mean, there is generally no corresponding closed-form
matrix average. Instead, verify positive definiteness and stationarity. Whitening
at $G$ turns its stationarity condition into
$B^{-1}\sum_b\log(G^{-1/2}C_bG^{-1/2})=0$; the Frobenius norm of this expression
is the metric gradient norm. Compute it again using base R.

```{r airm-reference}
inverse_root <- matrix_function(means$affine_invariant, function(x) x^(-0.5))
whitened_logs <- lapply(blocks, function(C) {
  matrix_function(inverse_root %*% C %*% inverse_root, log)
})
airm_stationarity <- norm(Reduce(`+`, whitened_logs) / length(blocks), "F")
smallest_eigenvalues <- vapply(means, function(G) {
  min(eigen(G, symmetric = TRUE)$values)
}, numeric(1))
stopifnot(log_reference_error < 1e-7, airm_stationarity < 1e-6,
          all(smallest_eigenvalues > 0),
          abs(airm_stationarity - fits$affine_invariant$gradient_norm) < 1e-8,
          all(vapply(names(fits), function(name) {
            riem.belongs(problems[[name]]$manifold, fits[[name]]$point)
          }, logical(1))))
data.frame(check = c("Relative log-Euclidean reference error",
                     "Independent affine-invariant stationarity",
                     "Smallest log-Euclidean eigenvalue",
                     "Smallest affine-invariant eigenvalue"),
           value = c(log_reference_error, airm_stationarity, smallest_eigenvalues))
```

## Compare the averages and block distances

The heatmaps share one color scale and label each covariance in squared
percentage log-return units. The arithmetic mean is included as a familiar
reference.

```{r covariance-heatmaps, fig.cap="The arithmetic and two geometric means use the same color scale; cell labels give covariances in squared percentage log-return units.", fig.alt="Three four-by-four heatmaps with numeric cell labels compare the arithmetic, log-Euclidean, and affine-invariant covariance means across DAX, SMI, CAC, and FTSE.", fig.width=9, fig.height=3.4}
panels <- c(list(Arithmetic = arithmetic),
             list(`Log-Euclidean` = means$log_euclidean,
                  `Affine-invariant` = means$affine_invariant))
limits <- range(unlist(panels))
palette <- hcl.colors(60, "YlOrRd")
color_breaks <- seq(limits[1], limits[2], length.out = length(palette) + 1L)
old_par <- par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
for (name in names(panels)) {
  image(seq_len(4), seq_len(4), panels[[name]], col = palette,
        zlim = limits, axes = FALSE, xlab = "", ylab = "", main = name, asp = 1)
  axis(1, seq_len(4), colnames(prices), cex.axis = 0.8)
  axis(2, seq_len(4), colnames(prices), cex.axis = 0.8, las = 2)
  values <- as.vector(panels[[name]])
  cell_colors <- palette[findInterval(values, color_breaks, all.inside = TRUE)]
  brightness <- drop(c(0.299, 0.587, 0.114) %*% col2rgb(cell_colors))
  text(rep(seq_len(4), times = 4), rep(seq_len(4), each = 4),
       labels = formatC(values, format = "f", digits = 2),
       col = ifelse(brightness > 145, "black", "white"), cex = 1)
}
par(old_par)
```

Distances between **consecutive blocks** reveal changes in covariance, while
distances from a mean summarize variation around a chosen center. These are
descriptive geometric distances, not tests of structural change.

```{r block-distances}
distance_to_mean <- sapply(names(problems), function(name) {
  vapply(tensor_blocks, function(C) {
    as.numeric(riem.dist(problems[[name]]$manifold, fits[[name]]$point, C))
  }, numeric(1))
})
consecutive_distance <- sapply(problems, function(problem) {
  vapply(seq_len(block_count - 1L), function(i) {
    as.numeric(riem.dist(problem$manifold, tensor_blocks[[i]], tensor_blocks[[i + 1L]]))
  }, numeric(1))
})
```

```{r distance-plots, fig.cap="Left: each block's distance from its corresponding geometric mean. Right: distances between consecutive blocks.", fig.alt="Two line charts compare log-Euclidean and affine-invariant distances, first from each block to its mean and then between consecutive covariance blocks."}
old_par <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
colors <- c("#21618C", "#C0652A")
matplot(seq_len(block_count), distance_to_mean, type = "b", pch = c(19, 17),
        lty = 1:2, col = colors, xlab = "120-observation block", ylab = "Distance to mean")
legend("topleft", c("Log-Euclidean", "Affine-invariant"), col = colors,
       lty = 1:2, pch = c(19, 17), bty = "n", cex = 0.75)
matplot(2:block_count, consecutive_distance, type = "b", pch = c(19, 17),
        lty = 1:2, col = colors, xlab = "Later block", ylab = "Distance from previous block")
par(old_par)
```

The block length and shrinkage affect these results. Blocks follow the stored
business-time order, rather than exact calendar semesters. Both geometric means
are SPD, but an averaging geometry is a modeling choice. The independent checks
establish solutions to the stated optimization problems. Assessing other block
lengths and shrinkage levels would help distinguish changes in the data from
effects of preprocessing.

CPU float64 is explicit throughout. These alternatives illustrate device
discovery and overrides without assuming accelerator hardware is available.
See [Applications and devices](applications-and-devices.html) for capability checks.

```{r device-options, eval=FALSE}
riem.optimize(problems$affine_invariant, initial, device = "auto")
riem.optimize(problems$affine_invariant, initial, device = "cpu")
riem.optimize(problems$affine_invariant, initial, device = "cuda:0")
```

```{r reproducibility-result, include=FALSE}
example_result <- c(log_reference_error = log_reference_error,
                    airm_stationarity = airm_stationarity,
                    lerm_objective = fits$log_euclidean$objective,
                    airm_objective = fits$affine_invariant$objective,
                    smallest_eigenvalues,
                    lerm_mean = as.vector(means$log_euclidean),
                    airm_mean = as.vector(means$affine_invariant))
names(example_result) <- make.unique(names(example_result))
stopifnot(all(is.finite(example_result)))
```
