---
title: "Robust regression on industrial measurements"
description: "Fit quadratic and Huber regression to stackloss data, inspect weights, and verify both optima with independent base R calculations."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
if (!torch::torch_is_installed()) {
  stop("This article needs the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(6103)
set.seed(6103)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.6, fig.align = "center", dpi = 120)
```

A few large residuals can strongly affect least-squares regression. A Huber
loss changes how the solver weighs those residuals while keeping the same
linear prediction function. Here we fit both choices to industrial
measurements and verify their answers independently.

Start with [the short robust-line example](example-robust-regression.html) if
residual-based problems are new to you. The [least-squares guide](../articles/second-order-and-leastsquares.html)
and [`riem.problem.leastsquares()`](../reference/riem.problem.leastsquares.html)
describe the interface used below.

## Data and predictor scaling

R's [stackloss data](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/stackloss.html)
record 21 operating days at a nitric-acid plant. The response measures ammonia
that escapes unabsorbed, in ten times percentage units. We use cooling-air
flow, water temperature, and acid concentration as predictors. The historical
acid column uses a coded scale; consult the linked help for its interpretation.
See [data sources and reproducibility](example-data-sources.html) for provenance.

```{r data}
observations <- datasets::stackloss
stopifnot(!anyNA(observations))
scaled_predictors <- scale(as.matrix(observations[, 1:3]))
design <- cbind(intercept = 1, scaled_predictors)
y <- observations$stack.loss
registered <- list(
  X = torch_tensor(design, dtype = torch_float64(), device = "cpu"),
  y = torch_tensor(y, dtype = torch_float64(), device = "cpu")
)
data.frame(observations = nrow(design), predictors = ncol(design) - 1)
```

We subtract each predictor's mean and divide by its sample standard deviation.
This improves numerical scaling but does not transform the response. An
intercept and three slopes form a point in four-dimensional Euclidean space.

## Objective: the same residual, two losses

For residual $r_i=X_i\beta-y_i$, ordinary least squares minimizes
$\sum_i r_i^2/2$. Huber regression minimizes $\sum_i\rho_\delta(r_i)$ with

\[
\rho_\delta(r)=
\begin{cases}
 r^2/2, & |r|\leq\delta,\\
 \delta(|r|-\delta/2), & |r|>\delta.
\end{cases}
\]

We choose **$\delta=2$ response units**, corresponding to 0.2 percentage points
in unabsorbed ammonia. This is an explicit illustrative modeling choice, not an
estimated noise scale. Changing it changes the fitted model.

```{r problem}
threshold <- 2
geometry <- manifold.euclidean(4)
make_problem <- function(loss) {
  riem.problem.leastsquares(
    geometry,
    residual = function(beta, data) data$X$matmul(beta) - data$y,
    data = registered, loss = loss, loss_scale = threshold
  )
}
```

The tensor data are registered so they move with the problem when the solver
selects a different device. Levenberg--Marquardt uses a positive-semidefinite
Gauss--Newton model; for Huber loss this includes residual weights and is not
the exact robust Hessian.

## Solve and report the two fits

```{r solve}
losses <- c("linear", "huber")
fits <- setNames(lapply(losses, function(loss) {
  riem.optimize(
    make_problem(loss),
    torch_zeros(4, dtype = torch_float64(), device = "cpu"),
    method = "levenberg_marquardt",
    control = list(max_iterations = 150, gradient_tolerance = 1e-7),
    device = "cpu", dtype = "float64"
  )
}), losses)
estimates <- t(vapply(fits, function(fit) as.numeric(fit$point), numeric(4)))
colnames(estimates) <- colnames(design)
round(estimates, 4)
data.frame(
  loss = losses,
  objective = vapply(fits, function(fit) fit$objective, numeric(1)),
  gradient_norm = vapply(fits, function(fit) fit$gradient_norm, numeric(1)),
  termination = vapply(fits, function(fit) fit$termination, character(1))
)
```

Slopes above are per one sample standard deviation of a predictor; the intercept
is at the mean predictor values. A `converged_gradient` result establishes the
requested first-order tolerance. If a run instead reaches `max_iterations`,
inspect its gradient before accepting it as converged. Raw objective values
from the two different losses are not comparable measures of model quality.

```{r fitted-residuals, echo=FALSE, fig.width=8, fig.height=4, fig.alt="Left: observed versus fitted stack loss for quadratic and Huber fits. Right: residuals versus fitted values, with dashed lines marking the positive and negative Huber threshold.", fig.cap="Both fits use the same predictors. Huber loss limits the influence of observations with large residuals."}
predicted <- design %*% t(estimates)
residuals <- sweep(predicted, 1, y, "-")
local({
  old <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.3, 0.7))
  on.exit(par(old))
  colors <- c(linear = "#446AA0", huber = "#267C74")
  limits <- range(c(y, predicted))
  plot(y, predicted[, "linear"], pch = 1, col = colors["linear"],
       xlim = limits, ylim = limits, xlab = "Observed stack loss",
       ylab = "Fitted stack loss", main = "Fitted responses")
  points(y, predicted[, "huber"], pch = 16, col = colors["huber"])
  abline(0, 1, lty = 3, col = "#7F8991")
  legend("topleft", c("Quadratic", "Huber"), col = colors,
         pch = c(1, 16), bty = "n", cex = 0.85)
  plot(predicted[, "linear"], residuals[, "linear"], pch = 1,
       col = colors["linear"], xlim = range(predicted), ylim = range(residuals),
       xlab = "Fitted stack loss", ylab = "Fitted minus observed",
       main = "Residuals")
  points(predicted[, "huber"], residuals[, "huber"], pch = 16,
         col = colors["huber"])
  abline(h = 0, col = "#7F8991", lty = 3)
  abline(h = c(-threshold, threshold), col = "#CB713C", lty = 2)
})
```

For Huber loss, the Gauss--Newton weight is
$w_i=\min(1,\delta/|r_i|)$, with weight one at a zero residual. Downweighting
does not delete a row or demonstrate that the observation is erroneous.

```{r robust-weights, fig.height=3.7, fig.alt="One bar per operating day shows its Huber residual weight, between zero and one; days with large residuals have smaller weights.", fig.cap="Weights are determined by the fitted residuals and the chosen threshold of two response units."}
weights <- pmin(1, threshold / pmax(abs(residuals[, "huber"]),
                                   .Machine$double.eps))
barplot(weights, names.arg = seq_along(weights), col = "#267C74",
        border = NA, ylim = c(0, 1.08), xlab = "Observation (operating day)",
        ylab = "Huber weight", main = "Which residuals receive less weight?",
        cex.names = 0.7)
abline(h = 1, lty = 3, col = "#7F8991")
```

## Independent verification

`lm()` solves the quadratic problem independently. For Huber loss, we implement
the objective and its gradient using base R and run `optim()` from the same
zero coefficients. Clipping residuals at the threshold gives the Huber score
$\psi_\delta(r)$, so the coefficient gradient is $X^\mathsf{T}\psi_\delta(r)$.

```{r verification}
quadratic_reference <- coef(lm(y ~ scaled_predictors))
huber_objective <- function(beta) {
  r <- drop(design %*% beta) - y
  sum(ifelse(abs(r) <= threshold, r^2 / 2,
             threshold * (abs(r) - threshold / 2)))
}
huber_gradient <- function(beta) {
  r <- drop(design %*% beta) - y
  drop(crossprod(design, pmax(-threshold, pmin(threshold, r))))
}
huber_reference <- optim(
  rep(0, 4), huber_objective, gr = huber_gradient, method = "BFGS",
  control = list(reltol = 1e-14, maxit = 1000)
)
coefficient_error <- c(
  linear = max(abs(estimates["linear", ] - quadratic_reference)),
  huber = max(abs(estimates["huber", ] - huber_reference$par))
)
independent_huber_gradient <- sqrt(sum(huber_gradient(estimates["huber", ])^2))
data.frame(loss = losses, maximum_coefficient_error = coefficient_error)
independent_huber_gradient
stopifnot(
  huber_reference$convergence == 0,
  all(coefficient_error < 1e-5),
  independent_huber_gradient < 1e-6,
  abs(huber_objective(estimates["huber", ]) - fits$huber$objective) < 1e-9,
  all(weights > 0 & weights <= 1),
  all(vapply(fits, function(fit) fit$gradient_norm < 1e-6, logical(1)))
)
```

The coefficient tolerance is `1e-5` because the independent solvers use different
stopping rules. A second seeded execution should reproduce the reported
coefficients and residual checks to absolute tolerance `1e-10` on the same
CPU/runtime.

## Interpretation and device options

This example illustrates an optimization choice. It does not establish which
loss predicts future plant behavior best; that would require a defensible
validation design and subject-matter assessment. Huber loss limits response
residual influence but does not by itself solve all leverage or model
misspecification problems.

The website runs on CPU float64. You can retain that override on a workstation
with GPUs, allow automatic selection, or explicitly request a compatible GPU:

```{r device-options, eval=FALSE}
initial <- torch_zeros(4, dtype = torch_float64(), device = "cpu")
riem.optimize(make_problem("huber"), initial, "levenberg_marquardt",
              device = "auto", dtype = "float64")
riem.optimize(make_problem("huber"), initial, "levenberg_marquardt",
              device = "cpu", dtype = "float64")
riem.optimize(make_problem("huber"), initial, "levenberg_marquardt",
              device = "cuda:0", dtype = "float64")
```

```{r reproducibility-result, include=FALSE}
example_result <- c(
  setNames(as.numeric(estimates), paste0("coefficient_", seq_len(8))),
  linear_objective = fits$linear$objective, huber_objective = fits$huber$objective,
  linear_gradient = fits$linear$gradient_norm,
  huber_gradient = fits$huber$gradient_norm,
  huber_reference_error = coefficient_error[["huber"]],
  minimum_weight = min(weights), independent_huber_gradient = independent_huber_gradient
)
stopifnot(all(is.finite(example_result)))
```
