---
title: "Align hand shapes without reflections"
description: "Align planar hand landmarks on the rotation manifold and check each solution against determinant-corrected SVD."
output: rmarkdown::html_document
resource_files:
  - data/riemann-hands.rds
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
if (!torch::torch_is_installed()) {
  stop("This article needs the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(7102)
set.seed(7102)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4.5, fig.align = "center", dpi = 120)
```

Two outlines can have the same shape but different positions, sizes, and
orientations. Centering and normalization remove position and positive scale;
a rotation can then align corresponding landmarks while preserving handedness.
This small example introduces optimization over rotation matrices, with an
independent closed-form solution for comparison.

Start with [getting started](getting-started.html) for the problem-and-solver
interface. The [geometry and metrics guide](geometry-and-metrics.html) explains
how the geometry determines the gradient and feasible search directions.

## Data: corresponding hand landmarks

The `hands` dataset from **Riemann 0.1.7** contains 40 configurations of four
people's left hands, with 56 corresponding planar landmarks per configuration.
For each person, the first six configurations run from maximally to minimally
spread fingers. We use these six configurations for person 1 and align five of
them to configuration 3. They are repeated poses of one person, not six
independent people.

The bundled [snapshot](data/riemann-hands.rds) is extracted from
[Riemann's pinned source](https://github.com/kisungyou/Riemann/blob/e847692576f8b250c8140e1343b145f72caa379f/data/hands.rda).
No installation of Riemann or download is needed. Landmark coordinates have no
documented physical unit; after normalization they are dimensionless. See
[data sources and reproducibility](example-data-sources.html) for provenance,
attribution, and extraction details.

```{r data}
hands <- readRDS("data/riemann-hands.rds")
stopifnot(identical(dim(hands$data), c(56L, 2L, 40L)),
          length(hands$person) == 40L, !anyNA(hands$data))
selected <- which(hands$person == 1)[1:6]

normalize_shape <- function(X) {
  centered <- sweep(X, 2, colMeans(X))
  centered / sqrt(sum(centered^2))
}
shapes <- lapply(selected, function(i) normalize_shape(hands$data[, , i]))
target <- shapes[[3]]
moving_indices <- c(1, 2, 4, 5, 6)

# Add known rotations solely to make the alignment step easy to see.
angles <- c(-65, -35, 25, 50, 80) * pi / 180
rotation <- function(angle) {
  matrix(c(cos(angle), sin(angle), -sin(angle), cos(angle)), 2, 2)
}
moving <- lapply(seq_along(moving_indices), function(j) {
  shapes[[moving_indices[j]]] %*% rotation(angles[j])
})
```

The added rotations are a demonstration transformation, **not a property of the
recorded poses**. The original variation in finger opening remains. Every shape
has centroid zero and Frobenius norm one; no reflection is applied.

## Objective and geometry

For a moving landmark matrix \(X\) and reference \(Y\), solve

\[
\min_{R\in\mathrm{SO}(2)} \frac12\lVert XR-Y\rVert_F^2,
\qquad R^\mathsf{T}R=I,\quad \det(R)=1.
\]

The rows of \(X\) are landmarks, so rotations multiply from the right.
[`manifold.rotation(2)`](../reference/manifold.rotation.html) represents the
orientation-preserving rotations with the embedded Frobenius metric. Restricting
the determinant to +1 prevents a left hand from being matched by a reflection.

```{r problem}
rotations <- manifold.rotation(2)
make_problem <- function(X, Y) {
  riem.problem(
    rotations,
    fn = function(R, data) (data$X$matmul(R) - data$Y)$square()$sum() / 2,
    data = list(
      X = torch_tensor(X, dtype = torch_float64(), device = "cpu"),
      Y = torch_tensor(Y, dtype = torch_float64(), device = "cpu")
    )
  )
}
```

Both objective tensors are registered through `data`, so the execution driver
can place them on the same device as the point. Automatic differentiation
provides the ambient gradient and the rotation geometry converts it to a
Riemannian gradient.

## Solve the five alignments

```{r solve}
initial <- torch_eye(2, dtype = torch_float64(), device = "cpu")
fits <- lapply(moving, function(X) {
  riem.optimize(
    make_problem(X, target), initial, method = "conjugate_gradient",
    control = list(max_iterations = 100, gradient_tolerance = 1e-7),
    device = "cpu"
  )
})
estimated_rotations <- lapply(fits, function(fit) as.matrix(fit$point))
aligned <- Map(function(X, R) X %*% R, moving, estimated_rotations)
before <- vapply(moving, function(X) sum((X - target)^2) / 2, numeric(1))
after <- vapply(fits, function(fit) fit$objective, numeric(1))
termination <- vapply(fits, function(fit) fit$termination, character(1))
data.frame(configuration = moving_indices, before, after, termination)
```

`converged_gradient` means the metric gradient norm passed the stopping
threshold. Exhausting the iteration budget is a different outcome; inspect
`termination` before interpreting a returned point as a converged solution.
Even at an optimum, different finger poses leave a nonzero alignment error.

## Plot the outlines before and after alignment

```{r alignment-plot, echo=FALSE, fig.alt="Two panels show five centered, normalized hand outlines. Added rotations spread their orientations before optimization; afterward their palms and fingers line up with the dark reference, while differences in finger opening remain.", fig.cap="Five poses of the same left hand, with demonstration rotations added on the left. Rotation alignment preserves the differences in finger opening visible on the right."}
colors <- c("#447A9C", "#CD713B", "#267C74", "#8F6BAA", "#BE647C")
limits <- range(unlist(c(moving, aligned, list(target)))) * 1.08
local({
  old_par <- par(mfrow = c(1, 2), mar = c(3.8, 3.8, 3, 0.7),
                 oma = c(2.4, 0, 0, 0), mgp = c(2.1, 0.65, 0))
  on.exit(par(old_par))
  draw <- function(outlines, title) {
    plot(NA, xlim = limits, ylim = limits, asp = 1,
         xlab = "Normalized coordinate 1", ylab = "Normalized coordinate 2",
         main = title, cex.lab = 0.85)
    for (j in seq_along(outlines)) {
      lines(rbind(outlines[[j]], outlines[[j]][1, ]), col = colors[j], lwd = 1.5)
    }
    lines(rbind(target, target[1, ]), col = "#243746", lwd = 2.5, lty = 2)
  }
  draw(moving, "Before: added rotations")
  draw(aligned, "After: rotations optimized")
  mtext("Colored outlines: configurations 1, 2, 4, 5, 6; dashed: reference 3",
        side = 1, outer = TRUE, line = 0.7, cex = 0.85)
})
```

## Independent verification

For this objective, a determinant-corrected SVD gives a global optimum.
If \(X^\mathsf{T}Y=U\Sigma V^\mathsf{T}\), then
\(R_*=U\operatorname{diag}(1,\det(UV^\mathsf{T}))V^\mathsf{T}\).
The determinant correction is needed because an unconstrained orthogonal
Procrustes solution may be a reflection. This reference uses base R's `svd()`.

```{r independent-verification}
svd_rotation <- function(X, Y) {
  decomposition <- svd(crossprod(X, Y))
  orientation <- det(decomposition$u %*% t(decomposition$v))
  decomposition$u %*% diag(c(1, sign(orientation))) %*% t(decomposition$v)
}
reference_rotations <- lapply(moving, svd_rotation, Y = target)
reference_objectives <- vapply(seq_along(moving), function(j) {
  sum((moving[[j]] %*% reference_rotations[[j]] - target)^2) / 2
}, numeric(1))
rotation_error <- max(vapply(seq_along(fits), function(j) {
  max(abs(estimated_rotations[[j]] - reference_rotations[[j]]))
}, numeric(1)))
orthogonality_error <- max(vapply(estimated_rotations, function(R) {
  max(abs(crossprod(R) - diag(2)))
}, numeric(1)))
determinant_error <- max(abs(vapply(estimated_rotations, det, numeric(1)) - 1))
objective_error <- max(abs(after - reference_objectives))

# Centering and normalization remove translation and positive uniform scale.
original <- hands$data[, , selected[1]]
transformed <- sweep(3.7 * original, 2, c(11, -7), "+")
normalization_error <- max(abs(normalize_shape(transformed) - shapes[[1]]))
checks <- c(rotation_error = rotation_error,
            orthogonality_error = orthogonality_error,
            determinant_error = determinant_error,
            objective_error = objective_error,
            normalization_error = normalization_error)
checks
stopifnot(all(termination == "converged_gradient"),
          rotation_error < 1e-6, orthogonality_error < 1e-10,
          determinant_error < 1e-10, objective_error < 1e-10,
          normalization_error < 1e-12, all(after <= before + 1e-12))
```

The rotation tolerance is looser than the objective tolerance because objective
errors are second order near a minimum. These checks also confirm preservation
of handedness and invariance of the preprocessing to translation and positive
scale.

## Interpretation and limits

This is alignment to a **fixed reference**, not an intrinsic mean on Kendall
shape space. It assumes that landmark rows correspond across configurations,
and keeps genuine differences in finger pose. It also differs from Riemann's
`wrap.landmark()` representation, which permits reflection equivalence. Here
the rotation manifold explicitly forbids reflections.

The SVD solution is the direct approach for this particular objective. The
optimization formulation becomes useful when an alignment objective includes
additional terms that no longer admit this closed-form solution. For another
short introduction with an independent matrix-factorization check, try
[principal-direction optimization](example-sphere-pca.html).

## Choose a device for your own run

The article uses explicit CPU float64 execution. For your own runs, change
`device` in [`riem.optimize()`](../reference/riem.optimize.html); registered
data move with the point. Such a small problem will usually not benefit from
GPU execution.

```{r device-options, eval=FALSE}
# device = "auto"    # discover a suitable device and preserve float64
# device = "cpu"     # override an available CUDA GPU
# device = "cuda:0"  # explicitly select the first CUDA GPU
```

```{r reproducibility-result, include=FALSE}
example_result <- c(before = before, after = after, checks)
stopifnot(is.numeric(example_result), all(is.finite(example_result)))
```
