---
title: "Find a principal direction on the sphere"
description: "A complete optimization example with iris data, a convergence plot, and an independent eigenvector check."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
if (!torch::torch_is_installed()) {
  stop("This article needs the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(2026)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.6, fig.align = "center", dpi = 120)
```

The first principal component is a unit vector that captures as much variation
as possible. That unit-length constraint makes the sphere a natural search
space. This example writes PCA as an optimization problem, solves it, and checks
the answer against R's `eigen()` function.

We use two measurements from the built-in `iris` data so the result is easy to
draw. The measurements are centered but kept in their original units.

## Build the objective

```{r data}
X <- scale(as.matrix(iris[, c("Sepal.Length", "Petal.Length")]),
           center = TRUE, scale = FALSE)
C <- crossprod(X) / nrow(X)

sphere <- manifold.sphere(2)
covariance <- torch_tensor(C, dtype = torch_float64(), device = "cpu")
problem <- riem.problem(
  sphere,
  function(q, data) -(q * data$matmul(q))$sum(),
  data = covariance
)
```

For a unit vector $q$, the projected variance is $q^\mathsf{T} Cq$. The solver
minimizes its negative. `riem.problem()` obtains the gradient by automatic
differentiation, and the geometry converts it to a tangent vector on the
sphere. Registering the covariance through `data` lets the execution driver
place it on the same device as the point.

## Solve and inspect the result

```{r solve}
initial <- torch_tensor(c(1, 0), dtype = torch_float64(), device = "cpu")
fit <- riem.optimize(
  problem, initial, method = "conjugate_gradient",
  control = list(max_iterations = 100, gradient_tolerance = 1e-7),
  device = "cpu"
)
fit
```

The examples on this site explicitly select CPU and float64 for reproducibility.
For automatic device discovery on your workstation, use `device = "auto"`.
You can always keep `device = "cpu"` to override an available CUDA GPU.

## Check against the eigensystem

The sign of a principal direction is arbitrary: $q$ and $-q$ describe the same
line. Compare their absolute inner product rather than their signed entries.

```{r reference}
reference <- eigen(C, symmetric = TRUE)
direction <- as.numeric(fit$point)
agreement <- abs(sum(direction * reference$vectors[, 1]))
captured <- -fit$objective / sum(diag(C))

data.frame(
  check = c("Unit-length error", "Direction agreement", "Variance captured"),
  value = c(abs(sum(direction^2) - 1), agreement, captured)
)
stopifnot(riem.belongs(sphere, fit$point), agreement > 1 - 1e-10)
```

The variance fraction is for these two centered measurements. It is not the
variance fraction for all four iris measurements or for standardized data.

```{r direction-plot, echo=FALSE, fig.alt="Centered iris measurements with the optimized first principal direction drawn through the origin.", fig.cap="The optimized direction follows the strongest variation in the two measurements."}
plot(X, pch = 19, col = grDevices::adjustcolor("#447A9C", alpha.f = 0.5),
     xlab = "Centered sepal length (cm)",
     ylab = "Centered petal length (cm)", asp = 1,
     main = "A principal direction from optimization")
abline(h = 0, v = 0, col = "#E2E8ED", lty = 3)
abline(a = 0, b = direction[2] / direction[1], col = "#CD713B", lwd = 3)
legend("topleft", c("Measurements", "Principal direction"),
       pch = c(19, NA), lty = c(NA, 1), lwd = c(NA, 3),
       col = c("#447A9C", "#CD713B"), bty = "n", cex = 0.85)
```

## Read the convergence history

The result retains the objective and metric gradient norm at each iteration.
Here the known optimum is minus the largest eigenvalue, so we can also measure
the objective gap directly.

```{r convergence, fig.alt="The objective gap decreases toward machine precision as conjugate-gradient iterations proceed.", fig.cap="The objective approaches the independently known optimum."}
iteration <- vapply(fit$history, function(record) record$iteration, numeric(1))
objective <- vapply(fit$history, function(record) record$objective, numeric(1))
gap <- pmax(objective + reference$values[1], .Machine$double.eps)
plot(iteration, gap, type = "l", log = "y", lwd = 2, col = "#267C74",
     xlab = "Iteration", ylab = "Objective gap (log scale)",
     main = "Progress toward the known optimum")
```

The plotted gap is clipped at machine precision because subtraction near the
optimum is affected by floating-point rounding. The solver's stopping criterion
uses the gradient norm. For several principal directions together, use
`manifold.grassmann()` and an orthonormal matrix-valued point.

Try [the penguin subspace application](example-penguin-subspace.html) to fit
two directions from public measurements and compare a finite-sum solver.
Browse all ten applications and introductions in the [example gallery](examples.html).
