---
title: "Fit a line with a robust loss"
description: "Compare ordinary least squares and Huber loss using the same residual function and verify both answers independently."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
if (!torch::torch_is_installed()) {
  stop("This article needs the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(2026)
set.seed(51)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.6, fig.align = "center", dpi = 120)
```

A least-squares problem separates the residual function from the choice of
loss. This makes it easy to compare a quadratic loss with a robust alternative.
We fit a line to a small synthetic data set containing three positive outliers.
Euclidean space is a manifold too, so the same problem and solver interface
applies without a geometric constraint.

## Generate observations and register the data

```{r observations}
set.seed(51)
x <- seq(-2, 2, length.out = 41)
y <- 1 + 1.5 * x + rnorm(length(x), sd = 0.15)
outliers <- c(31, 35, 39)
y[outliers] <- y[outliers] + 3

design <- cbind(intercept = 1, slope = x)
registered <- list(
  X = torch_tensor(design, dtype = torch_float64(), device = "cpu"),
  y = torch_tensor(y, dtype = torch_float64(), device = "cpu")
)
```

The point has two coordinates: the intercept and the slope. The residual
callback returns one value for each observation, leaving the constructor to
apply the chosen loss.

```{r problem}
coefficients <- manifold.euclidean(2)
make_problem <- function(loss) {
  riem.problem.leastsquares(
    coefficients,
    residual = function(beta, data) data$X$matmul(beta) - data$y,
    data = registered,
    loss = loss,
    loss_scale = 0.25
  )
}
```

With `loss = "linear"`, the objective is one half of the sum of squared
residuals. With `loss = "huber"`, each residual has a quadratic loss up to an
absolute value of `loss_scale`, and a linear loss beyond it. The threshold is
in the units of the response; the value `0.25` is a choice for this simulated
example, not an automatically estimated noise scale.

## Solve both problems

```{r solve}
losses <- c("linear", "huber")
fits <- setNames(lapply(losses, function(loss) {
  riem.optimize(
    make_problem(loss),
    torch_zeros(2, dtype = torch_float64(), device = "cpu"),
    method = "levenberg_marquardt",
    control = list(max_iterations = 100, gradient_tolerance = 1e-7),
    device = "cpu"
  )
}), losses)

estimates <- t(vapply(fits, function(fit) as.numeric(fit$point), numeric(2)))
colnames(estimates) <- c("Intercept", "Slope")
round(rbind(Generating_line = c(1, 1.5), estimates), 4)
```

Levenberg--Marquardt uses a positive-semidefinite Gauss--Newton approximation.
For robust losses it reweights individual residuals; that approximation is
not the full Hessian of the robust objective. Automatic differentiation supplies
the residual derivatives here.

```{r fits-plot, echo=FALSE, fig.alt="Observed responses with three marked outliers, and generating, least-squares, and Huber regression lines.", fig.cap="Huber loss reduces the effect of the three added outliers in this example."}
plot(x, y, pch = 19, col = "#718494", xlab = "Predictor", ylab = "Response",
     ylim = range(c(y, -2.7, 7.5)), main = "The same residuals, two losses")
points(x[outliers], y[outliers], pch = 21, bg = "#CD713B", col = "#9C4D26",
       cex = 1.3)
abline(a = 1, b = 1.5, col = "#596775", lty = 3, lwd = 2)
abline(a = estimates["linear", 1], b = estimates["linear", 2],
       col = "#446AA0", lwd = 2)
abline(a = estimates["huber", 1], b = estimates["huber", 2],
       col = "#267C74", lwd = 2)
legend("topleft", c("Generating line", "Least squares", "Huber"),
       col = c("#596775", "#446AA0", "#267C74"), lty = c(3, 1, 1),
       lwd = 2, bty = "n", cex = 0.85)
```

## Verify against independent R calculations

For ordinary least squares, `lm.fit()` gives a direct reference. For Huber loss,
write the scalar objective in base R and minimize it with `optim()`. The latter
is a numerical cross-check rather than a closed-form solution.

```{r reference}
least_squares_reference <- lm.fit(design, y)$coefficients
huber_objective <- function(beta) {
  residual <- drop(design %*% beta) - y
  threshold <- 0.25
  sum(ifelse(abs(residual) <= threshold,
             residual^2 / 2,
             threshold * (abs(residual) - threshold / 2)))
}
huber_reference <- optim(c(0, 0), huber_objective, method = "BFGS",
                         control = list(reltol = 1e-12))

comparison <- data.frame(
  loss = losses,
  maximum_coefficient_error = c(
    max(abs(estimates["linear", ] - least_squares_reference)),
    max(abs(estimates["huber", ] - huber_reference$par))
  ),
  gradient_norm = vapply(fits, function(fit) fit$gradient_norm, numeric(1)),
  termination = vapply(fits, function(fit) fit$termination, character(1))
)
comparison
stopifnot(huber_reference$convergence == 0,
          all(comparison$maximum_coefficient_error < 1e-5))
```

The two loss functions have different scales and meanings, so their raw
objective values should not be used to rank the fits. Compare the coefficients,
the residuals, and performance on data appropriate to your application.

`riem.problem.leastsquares()` also supports `"soft_l1"` and `"cauchy"` losses.
Nonquadratic losses support diagonal residual weights. For automatic device
selection, replace the explicit CPU override in the solve with
`device = "auto"`; registered data travel with the problem.

Continue with [industrial stack-loss measurements](example-stackloss-regression.html)
for a public-data application, or add explicit modeling assumptions in the
[constrained tree-volume example](example-tree-constraints.html).
