---
title: "Find a two-dimensional penguin subspace"
description: "Optimize a Grassmann subspace for Palmer Penguins measurements, compare with PCA, and try a finite-sum solver."
output: rmarkdown::html_document
resource_files:
  - data/penguins.csv
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
if (!torch::torch_is_installed()) {
  stop("This article needs the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(6101)
set.seed(6101)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.8, fig.align = "center", dpi = 120)
```

How can four body measurements be summarized in two coordinates? PCA answers
this by finding a plane with small reconstruction error. The plane is a point
on a Grassmann manifold: different orthonormal bases describe the same point.
Here we optimize that plane directly and compare the result with `prcomp()`.

This builds on [the sphere PCA introduction](example-sphere-pca.html).
For the geometry, see [geometry and metrics](../articles/geometry-and-metrics.html)
and [`manifold.grassmann()`](../reference/manifold.stiefel.html).

## Data: four measurements, one common scale

The [Palmer Penguins data](https://allisonhorst.github.io/palmerpenguins/) were
collected by Kristen Gorman and Palmer Station LTER. This site bundles the
CC0 `penguins` data from `palmerpenguins` 0.1.1, so building this article requires
no download. See [data sources and reproducibility](example-data-sources.html)
for attribution and the snapshot checksum.

```{r data}
penguins <- read.csv("data/penguins.csv", stringsAsFactors = FALSE)
measurements <- c("bill_length_mm", "bill_depth_mm",
                  "flipper_length_mm", "body_mass_g")
complete <- complete.cases(penguins[, measurements])
X <- scale(as.matrix(penguins[complete, measurements]))
species <- factor(penguins$species[complete])
data.frame(total = nrow(penguins), retained = nrow(X),
           missing_measurements = sum(!complete))
```

Bill and flipper measurements are in millimeters; mass is in grams. Centering
and dividing each column by its sample standard deviation gives the four
variables equal variance before optimization. Only missing body measurements
cause row removal; missing sex values do not. Species labels color the final
plot but are never inputs to the objective.

## Objective and geometry

Let $X$ contain the standardized observations, and let $Q$ be a $4\times2$
matrix with orthonormal columns. We minimize

\[
  f(Q) = \frac{1}{2n}\|X-XQQ^\mathsf{T}\|_F^2.
\]

Multiplying $Q$ by any orthogonal $2\times2$ matrix leaves its projector
$QQ^\mathsf{T}$, and therefore the objective, unchanged. This is why we use
[`manifold.grassmann(4, 2)`](../reference/manifold.stiefel.html).

```{r problem}
geometry <- manifold.grassmann(4, 2)
observations <- torch_tensor(X, dtype = torch_float64(), device = "cpu")
problem <- riem.problem(
  geometry,
  function(Q, data) {
    residual <- data - data$matmul(Q)$matmul(Q$t())
    (residual^2)$sum() / (2 * data$size(1))
  },
  data = observations
)
initial <- riem.random(geometry, device = "cpu", dtype = "float64")
```

Registering tensors through `data` lets the solver move them together with the
point when the execution device changes. Automatic differentiation computes the
ambient derivative, and the geometry supplies the tangent gradient.

## Solve and inspect termination

```{r solve}
fit <- riem.optimize(
  problem, initial, method = "conjugate_gradient",
  control = list(max_iterations = 200, gradient_tolerance = 1e-7),
  device = "cpu", dtype = "float64"
)
fit
Q <- as.matrix(fit$point)
scores <- X %*% Q
```

`converged_gradient` means the metric gradient met the requested tolerance.
`max_iterations` means the iteration allowance was exhausted, which alone does
not establish convergence. Keep both the termination reason and gradient norm
when comparing solvers.

```{r penguin-projection, echo=FALSE, fig.alt="Palmer Penguins projected onto the optimized two-dimensional plane, with points colored by Adelie, Chinstrap, and Gentoo species.", fig.cap="Species are used only to color the projection. The optimized basis can rotate without changing the plane."}
palette <- c(Adelie = "#CB713C", Chinstrap = "#78609E", Gentoo = "#267C74")
plot(scores, col = grDevices::adjustcolor(palette[species], alpha.f = 0.7),
     pch = 19, xlab = "Subspace coordinate 1", ylab = "Subspace coordinate 2",
     asp = 1, main = "A plane fitted to four body measurements")
legend("topright", legend = levels(species), col = palette[levels(species)],
       pch = 19, bty = "n", cex = 0.85)
```

The two displayed axes are coordinates in the fitted plane; they need not
equal the ordered first and second principal-component axes. This is an
exploratory representation, not a trained species classifier.

## A short finite-sum comparison

The same objective is the mean of one reconstruction loss per penguin.
[`riem.problem.finitesum()`](../reference/riem.problem.finitesum.html) can
evaluate a selected minibatch in one tensor operation through `batch_fn`.

```{r finite-sum}
finite_problem <- riem.problem.finitesum(
  geometry,
  function(Q, i, data) {
    observation <- data[i, ]
    residual <- observation - observation$matmul(Q)$matmul(Q$t())
    (residual^2)$sum() / 2
  },
  n = nrow(X),
  batch_fn = function(Q, indices, data) {
    batch <- data[indices, , drop = FALSE]
    residual <- batch - batch$matmul(Q)$matmul(Q$t())
    (residual^2)$sum(dim = 2) / 2
  },
  data = observations
)
torch_manual_seed(6102)
svrg_fit <- riem.optimize(
  finite_problem, initial, method = "svrg",
  control = list(max_iterations = 120, batch_size = 32, step_size = 0.15,
                 epoch_length = 20, full_evaluation_every = 20,
                 gradient_tolerance = 1e-7),
  device = "cpu", dtype = "float64"
)
fits <- list(conjugate_gradient = fit, svrg = svrg_fit)
data.frame(
  method = names(fits),
  objective = vapply(fits, function(x) x$objective, numeric(1)),
  full_gradient_norm = vapply(fits, function(x) x$gradient_norm, numeric(1)),
  termination = vapply(fits, function(x) x$termination, character(1))
)
```

SVRG uses a full gradient at a snapshot plus transported minibatch corrections.
The Grassmann frame representation supports the local logarithm needed to
connect each snapshot to the current point. The small steps here stay within
that local domain; the path is not defined at a Grassmann cut locus. Unsupported
geometries or paths fail explicitly. See [variance reduction and solver
controls](../articles/variance-reduction.html) for the contract.

This deliberately short run is a demonstration, not a speed benchmark. Its
final gradient is evaluated on all observations; the minibatch gradient does
not certify stationarity. A small data matrix usually favors a deterministic
method. More iterations or different steps may be needed for SVRG to satisfy
the same stopping tolerance.

## Independent verification

Compare projectors, because comparing signed basis entries would penalize
equivalent planes. `prcomp()` uses a separate singular-value decomposition.
The fraction of standardized variance retained is
$\|XQ\|_F^2/\|X\|_F^2$.

```{r verification}
reference <- prcomp(X, center = FALSE, scale. = FALSE)
reference_Q <- reference$rotation[, 1:2, drop = FALSE]
projector_error <- max(abs(tcrossprod(Q) - tcrossprod(reference_Q)))
captured <- sum(scores^2) / sum(X^2)
reference_captured <- sum(reference$sdev[1:2]^2) / sum(reference$sdev^2)

angle <- pi / 6
rotation <- matrix(c(cos(angle), sin(angle), -sin(angle), cos(angle)), 2, 2)
rotated_Q <- Q %*% rotation
rotation_error <- max(abs(tcrossprod(Q) - tcrossprod(rotated_Q)))
reconstruction_value <- function(basis) {
  sum((X - X %*% tcrossprod(basis))^2) / (2 * nrow(X))
}
finite_value <- as.numeric(riem.evaluate(finite_problem, fit$point,
                                        gradient = FALSE)$value)
initial_value <- as.numeric(riem.evaluate(problem, initial,
                                         gradient = FALSE)$value)
data.frame(
  projector_error, captured, reference_captured, rotation_error,
  svrg_objective_gap = svrg_fit$objective - fit$objective
)
stopifnot(
  all(riem.belongs(geometry, fit$point)),
  projector_error < 1e-6,
  abs(captured - reference_captured) < 1e-10,
  rotation_error < 1e-12,
  abs(reconstruction_value(rotated_Q) - fit$objective) < 1e-10,
  abs(finite_value - fit$objective) < 1e-10,
  fit$gradient_norm < 1e-6,
  is.finite(svrg_fit$gradient_norm),
  svrg_fit$objective < initial_value
)
```

These tolerances check the plane, the objective normalization, and basis
invariance separately. A second seeded execution should reproduce the reported
numeric checks to absolute tolerance `1e-10` on the same CPU/runtime.

## Device selection and interpretation

This article explicitly uses CPU float64. In your own session, the same
registered-data problem can select a suitable device automatically or honor
an explicit override:

```{r device-options, eval=FALSE}
riem.optimize(problem, initial, device = "auto", dtype = "float64")
riem.optimize(problem, initial, device = "cpu", dtype = "float64")
riem.optimize(problem, initial, device = "cuda:0", dtype = "float64")
```

The CUDA call requires an available compatible GPU. Results describe the
complete observed measurements after standardization; different scaling or
missing-data assumptions define a different analysis. The package supplies
the optimization tools rather than a separate statistical PCA estimator.

```{r reproducibility-result, include=FALSE}
example_result <- c(
  objective = fit$objective, gradient_norm = fit$gradient_norm,
  projector_error = projector_error, variance_fraction = captured,
  rotation_error = rotation_error, finite_value = finite_value,
  svrg_objective = svrg_fit$objective, svrg_gradient_norm = svrg_fit$gradient_norm,
  setNames(as.numeric(tcrossprod(Q)), paste0("projector_", seq_len(16)))
)
stopifnot(all(is.finite(example_result)))
```
