---
title: "Example gallery"
description: "Ten complete optimization examples, from first steps to public-data applications, constraints, and sparse solutions."
output: rmarkdown::html_document
resource_files:
  - data/example-runtimes.csv
---

Start with a foundation example to learn the problem interface, or choose a
public-data application close to your work. Every page contains runnable code,
figures, and checks of the numerical result. The six applications use data
already included with R or bundled in this website; no dataset download is
needed when building the documentation.

```{r setup, include=FALSE}
timings <- read.csv("data/example-runtimes.csv", stringsAsFactors = FALSE)
runtime_for <- function(article) {
  i <- match(article, timings$article)
  stopifnot(!anyNA(i), all(is.finite(timings$render_seconds[i])))
  sprintf("%.1f s", timings$render_seconds[i])
}
show_examples <- function(article, application, dataset, geometry, level) {
  knitr::kable(data.frame(
    Application = sprintf("[%s](%s.html)", application, article),
    Dataset = dataset, Geometry = geometry, Level = level,
    `CPU render` = runtime_for(article), check.names = FALSE
  ), row.names = FALSE)
}
```

## Foundations

These short introductions isolate the optimization ideas. The sphere example
uses two iris measurements; the other three construct small synthetic problems
with known reference answers.

```{r foundations, echo=FALSE, results='asis'}
show_examples(
  c("example-sphere-pca", "example-robust-regression", "example-spd-mean",
    "example-matrix-completion"),
  c("Find a principal direction", "Compare quadratic and Huber loss",
    "Compute a geometric midpoint", "Complete a low-rank matrix"),
  c("Iris", "Synthetic regression", "Two synthetic matrices", "Synthetic rank-two matrix"),
  c("Sphere", "Euclidean", "Positive-definite matrices", "Compact fixed rank"),
  c("Introductory", "Introductory", "Intermediate", "Intermediate")
)
```

## Public-data applications

Each application starts with the observations and explains how the objective
relates to the question being asked. The independent checks are available after
the main workflow, so you can follow the data analysis before studying its
numerical details.

```{r applications, echo=FALSE, results='asis'}
show_examples(
  c("example-penguin-subspace", "example-stackloss-regression",
    "example-airquality-completion", "example-covariance-means"),
  c("Find a penguin subspace", "Fit robust industrial regression",
    "Reconstruct missing measurements", "Average observed covariances"),
  c("Palmer Penguins", "Stack loss", "Air quality", "European stock indices"),
  c("Grassmann", "Euclidean", "Compact fixed rank", "SPD: log-Euclidean and affine-invariant"),
  c("Intermediate", "Introductory", "Intermediate", "Intermediate")
)
```

## Constraints and sparsity

Euclidean space also fits the manifold interface. These examples use it to
make equality constraints, inequalities, and an exact proximal operator easy
to understand and verify.

```{r constraints, echo=FALSE, results='asis'}
show_examples(
  c("example-tree-constraints", "example-mtcars-proximal"),
  c("Constrain a tree-volume model", "Trace a sparse regression path"),
  c("Black cherry trees", "Motor Trend cars"),
  c("Euclidean with equality/inequality constraints", "Euclidean with L1 penalty"),
  c("Intermediate", "Intermediate")
)
```

## Reading the results

The CPU times are measured complete article render times, not solver-only
benchmarks or speed guarantees. They include data preparation, independent
checks, figures, and HTML rendering in fresh R sessions. Values are the mean
of two serial runs on the reference machine: macOS arm64, R 4.5.2, torch 0.17.0,
CPU float64, one torch thread. The
[timing table](data/example-runtimes.csv) contains both measurements.

Each public-data example reports its termination reason and the residuals
appropriate to the method. A small training loss is not proof of accurate
missing-value predictions, and a stopped iteration budget is not convergence.
The articles discuss these distinctions using their actual results.

Use `device = "auto"` to request automatic device selection or `device = "cpu"`
to force CPU execution, including on a GPU workstation. An explicit CUDA device
must be available and support the requested precision. See the
[device guide](applications-and-devices.html).

For citations, units, missing values, the penguin CSV, and the reproducibility
procedure, see [Data sources and reproducibility](example-data-sources.html).
