---
title: "Reconstruct missing environmental measurements"
description: "Fit compact low-rank factors to New York air-quality measurements, with a leakage-free holdout and a simple imputation baseline."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(riemtorch)
library(torch)
if (!torch_is_installed()) {
  stop("This worked example requires the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(2026)
set.seed(2026)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4, fig.align = "center", dpi = 120)
```

Environmental measurements often contain missing entries and variables with
different units. This example treats four measurements as a partially observed
matrix and fits a rank-two approximation. A small random subset of the recorded
values is withheld so that prediction can be evaluated honestly.

Start with [the synthetic matrix-completion example](example-matrix-completion.html)
for the basic workflow. The [compact representation guide](compact-lowrank.html)
explains `manifold.fixedrank()` and `riem.materialize()`. Here, the data are noisy
and their true rank is unknown, so perfect reconstruction is not the objective.

## Data and a reproducible holdout

R's [`airquality` data](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/airquality.html)
contain 153 daily observations from New York in May--September 1973. We use ozone
(ppb), solar radiation (Langleys), wind (mph), and maximum temperature (degrees F).
Month and day are omitted from this matrix model. Data provenance and reuse
information appear on [Data sources and reproducibility](example-data-sources.html).

Withhold 10% of the observed entries that belong to rows initially containing
at least three measurements. Sample these candidates in a seeded random order,
accepting an entry only when its row will retain at least two training entries.
This preserves minimal information about every row. The holdout is separate
from genuinely missing values, for which no prediction error can be measured.

```{r split}
variables <- c("Ozone", "Solar.R", "Wind", "Temp")
measurements <- as.matrix(datasets::airquality[, variables])
observed <- !is.na(measurements)
training <- observed
held_out <- matrix(FALSE, nrow(measurements), ncol(measurements))
eligible <- which(observed & (rowSums(observed) >= 3), arr.ind = TRUE)
target <- floor(0.10 * nrow(eligible))
for (j in sample(seq_len(nrow(eligible)))) {
  row <- eligible[j, 1]
  column <- eligible[j, 2]
  if (sum(training[row, ]) > 2) {
    training[row, column] <- FALSE
    held_out[row, column] <- TRUE
  }
  if (sum(held_out) == target) break
}
stopifnot(sum(held_out) == target, all(rowSums(training) >= 2),
          !any(training & held_out), all((training | held_out) == observed))
data.frame(training_entries = sum(training), withheld_entries = sum(held_out),
           originally_missing = sum(!observed))
```

Estimate each column's mean and standard deviation from **training entries
only**. The centered, scaled training matrix is the input to optimization.
Held-out measurements enter only the later assessment.

```{r scale-training}
training_values <- measurements
training_values[!training] <- NA_real_
center <- colMeans(training_values, na.rm = TRUE)
spread <- apply(training_values, 2, sd, na.rm = TRUE)
stopifnot(all(is.finite(spread)), all(spread > 0))
standardized_training <- sweep(sweep(training_values, 2, center, "-"),
                               2, spread, "/")
indices <- which(training, arr.ind = TRUE)
stopifnot(!anyNA(standardized_training[training]),
          all(is.na(standardized_training[!training])))
```

## Objective and compact geometry

Write the approximation as $X = USV^\mathsf{T}$, with orthonormal columns in
$U$ and $V$ and a two-by-two core $S$. For training entries $\Omega$, minimize

\[
 f(X) = \frac{1}{2|\Omega|}\sum_{(i,j)\in\Omega}(X_{ij}-Z_{ij})^2
       + \frac{\lambda}{2np}\lVert X\rVert_F^2,\qquad \lambda=0.01.
\]

The penalty discourages large unobserved predictions. Because the outer factors
are orthonormal, $\lVert X\rVert_F=\lVert S\rVert_F$: the loss needs only selected
entries and the small core. The rank and penalty are fixed for this illustration;
the holdout is not used to choose them or select a starting point.

```{r problem}
n <- nrow(measurements)
p <- ncol(measurements)
lambda <- 0.01
M <- manifold.fixedrank(n, p, 2, representation = "svd")
problem <- riem.problem(
  M,
  function(x, data) {
    residual <- riem.materialize(M, x, data$rows, data$columns) - data$values
    residual$square()$mean() / 2 +
      x$S$square()$sum() * data$lambda / (2 * data$n * data$p)
  },
  data = list(rows = indices[, 1], columns = indices[, 2],
              values = torch_tensor(standardized_training[training],
                                    dtype = torch_float64(), device = "cpu"),
              lambda = lambda, n = n, p = p)
)
```

Registering tensors through `data` allows the solver to move them with the
optimization point. Initialize by filling unavailable standardized entries with
zero (the training column mean) and retaining the first two singular components.
This is a starting point, not a final imputation: optimization then fits only the
actual training observations. No holdout values enter the initialization.

```{r solve}
filled_training <- standardized_training
filled_training[!training] <- 0
decomposition <- svd(filled_training, nu = 2, nv = 2)
initial <- lapply(list(U = decomposition$u,
                       S = diag(decomposition$d[1:2]), V = decomposition$v),
                  torch_tensor, dtype = torch_float64(), device = "cpu")
fit <- riem.optimize(problem, initial, method = "lbfgs", device = "cpu",
                     control = list(max_iterations = 120, gradient_tolerance = 1e-6))
data.frame(objective = fit$objective, gradient_norm = fit$gradient_norm,
           iterations = fit$iterations, termination = fit$termination)
```

A maximum-iteration termination is a budget limit, not a convergence
certificate. Even a small gradient does not establish a global optimum for this
nonconvex problem. A real analysis should also examine other ranks and starts.

## Compare predictions with column means

Materialize the fitted matrix once for plots and assessment. Column-mean
imputation predicts zero on the standardized scale. We report the aggregate
standardized RMSE and separate original-unit errors; adding errors measured in
different physical units would be misleading.

```{r assessment}
reconstructed <- as.matrix(riem.materialize(M, fit$point))
prediction <- sweep(sweep(reconstructed, 2, spread, "*"), 2, center, "+")
baseline <- matrix(rep(center, each = n), nrow = n)
standardized_truth <- sweep(sweep(measurements, 2, center, "-"), 2, spread, "/")
standardized_rmse <- c(
  rank_two = sqrt(mean((reconstructed[held_out] - standardized_truth[held_out])^2)),
  column_mean = sqrt(mean(standardized_truth[held_out]^2))
)
standardized_rmse
original_rmse <- data.frame(
  variable = variables, units = c("ppb", "Langleys", "mph", "degrees F"),
  withheld = colSums(held_out),
  rank_two = vapply(seq_len(p), function(j) {
    keep <- held_out[, j]
    sqrt(mean((prediction[keep, j] - measurements[keep, j])^2))
  }, numeric(1)),
  column_mean = vapply(seq_len(p), function(j) {
    keep <- held_out[, j]
    sqrt(mean((baseline[keep, j] - measurements[keep, j])^2))
  }, numeric(1))
)
original_rmse
```

For this seeded split, the rank-two fit improves ozone prediction but has a
higher aggregate standardized error than column-mean imputation. A geometric
model can capture useful relationships without improving every variable or the
overall prediction score.

```{r patterns, fig.cap="Left: training, withheld, and genuinely missing entries. Right: the fitted standardized rank-two matrix.", fig.alt="A categorical heatmap separates observed training entries, deliberately withheld measurements, and original missing values for 153 days and four variables; a second heatmap shows the fitted standardized values."}
status <- matrix(0, n, p)
status[training] <- 1
status[held_out] <- 2
old_par <- par(mfrow = c(1, 2), mar = c(5, 4, 3, 1))
image(seq_len(p), seq_len(n), t(status), breaks = c(-0.5, 0.5, 1.5, 2.5),
      col = c("#E5E7EB", "#21618C", "#D68910"), axes = FALSE,
      xlab = "", ylab = "Day index", main = "Which entries are available?")
axis(1, seq_len(p), variables)
axis(2)
legend("bottom", inset = c(0, -0.32), xpd = NA, horiz = TRUE, bty = "n",
       legend = c("Missing", "Training", "Withheld"), cex = 0.75,
       fill = c("#E5E7EB", "#21618C", "#D68910"))
image(seq_len(p), seq_len(n), t(reconstructed), col = hcl.colors(60, "Blue-Red 3"),
      axes = FALSE, xlab = "", ylab = "Day index", main = "Reconstructed (standardized)")
axis(1, seq_len(p), variables)
axis(2)
par(old_par)
```

```{r heldout-plot, fig.width=6, fig.height=4.5, fig.cap="Predictions are assessed only where recorded measurements were withheld.", fig.alt="Scatterplot of withheld standardized measurements against rank-two predictions with a diagonal reference line and colors identifying the four variables."}
plot(standardized_truth[held_out], reconstructed[held_out],
     col = c("#21618C", "#B9770E", "#148F77", "#884EA0")[col(held_out)[held_out]],
     pch = 19, xlab = "Withheld measurement (standardized)",
     ylab = "Prediction (standardized)")
abline(0, 1, lty = 2, col = "grey40")
legend("topleft", variables, col = c("#21618C", "#B9770E", "#148F77", "#884EA0"),
       pch = 19, bty = "n", cex = 0.8)
```

## Independent numerical checks

Recalculate the objective with an ordinary dense R matrix, including the
Frobenius penalty. It must agree with the compact implementation. Verify that
the point remains on its manifold and that all predictions are finite. These
checks do not require the fitted method to outperform column means.

```{r verification}
dense_objective <- mean((reconstructed[training] -
                          standardized_training[training])^2) / 2 +
  lambda * sum(reconstructed^2) / (2 * n * p)
objective_error <- abs(dense_objective - fit$objective)
stopifnot(riem.belongs(M, fit$point), all(is.finite(reconstructed)),
          all(colSums(held_out) > 0), objective_error < 1e-10,
          abs(sum(reconstructed^2) - as.numeric(fit$point$S$square()$sum())) < 1e-8)
data.frame(dense_objective = dense_objective,
           compact_objective = fit$objective, absolute_difference = objective_error)
```

The random holdout approximates missing-at-random prediction among recorded
entries, subject to the row-coverage restriction. Real missingness may follow a
different mechanism. Time dependence, nonlinear relationships, and nonnegativity
are not modeled; some fitted physical measurements can therefore be implausible.
The baseline and held-out errors make these limitations visible.

This article explicitly forces CPU float64 for reproducibility. To use another
device, keep the same problem and change the argument below. Automatic selection
preserves precision; an unavailable explicitly requested GPU causes an error.
See [Applications and devices](applications-and-devices.html).

```{r device-options, eval=FALSE}
# Automatic discovery, or an explicit override:
riem.optimize(problem, initial, method = "lbfgs", device = "auto")
riem.optimize(problem, initial, method = "lbfgs", device = "cpu")
riem.optimize(problem, initial, method = "lbfgs", device = "cuda:0")
```

```{r reproducibility-result, include=FALSE}
example_result <- c(objective = fit$objective, gradient_norm = fit$gradient_norm,
                    standardized_rmse, as.vector(as.matrix(original_rmse[, 4:5])),
                    dense_objective_error = objective_error)
names(example_result) <- make.unique(ifelse(names(example_result) == "",
                                           "original_unit_rmse", names(example_result)))
stopifnot(all(is.finite(example_result)))
```
