---
title: "Fit a constrained tree-volume model"
description: "Use equality and inequality constraints for a log-volume model, then verify the solution by eliminating one slope."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
if (!torch::torch_is_installed()) {
  stop("This article needs the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(316)
set.seed(316)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.6, fig.align = "center", dpi = 120)
```

Suppose a volume model should increase with both diameter and height, and
multiplying both measurements by a common factor should multiply predicted
volume by its cube. These are **modeling assumptions**, not consequences of
fitting the data. They lead to a small regression problem with both an equality
and inequalities. Euclidean coefficients let us inspect the constraint solver
without introducing a geometric representation at the same time.

Start with [the regression introduction](example-robust-regression.html) if the
problem-and-solver interface is new to you. The
[constraints and nonsmooth guide](constrained-nonsmooth.html) explains the
solver contract in more detail.

## Data: diameter, height, and timber volume

R's [`trees` data](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/trees.html)
contain 31 black cherry trees. The misleadingly named `Girth` column records
**diameter in inches**, not circumference. Height is in feet and timber volume
in cubic feet. All three variables are positive and complete. No download is
needed; see [data sources and reproducibility](example-data-sources.html).

```{r data}
trees <- datasets::trees
stopifnot(nrow(trees) == 31, !anyNA(trees), all(as.matrix(trees) > 0))
log_predictors <- log(as.matrix(trees[, c("Girth", "Height")]))
log_volume <- log(trees$Volume)
predictor_center <- colMeans(log_predictors)
response_center <- mean(log_volume)
X <- cbind(intercept = 1, sweep(log_predictors, 2, predictor_center))
y <- log_volume - response_center

registered <- list(
  X = torch_tensor(X, dtype = torch_float64(), device = "cpu"),
  y = torch_tensor(y, dtype = torch_float64(), device = "cpu")
)
```

Centering the log predictors and response improves numerical conditioning and
changes only the intercept. We do not standardize the predictors, so the slope
constraints keep their original meaning.

## Objective and constraints

Write the model in the original coordinates as

\[
\log V_i \approx a + b_D\log D_i + b_H\log H_i,
\qquad b_D+b_H=3,\quad b_D\geq0,\quad b_H\geq0.
\]

We minimize the mean squared log residual divided by two. In the constructor,
equalities have the form \(h(\beta)=0\) and inequalities have the form
\(g(\beta)\leq0\). Our optimization point contains the centered intercept
followed by the two slopes. Registered data are passed to every callback, even
when a particular callback needs only the coefficients.

```{r problem}
problem <- riem.problem.constrained(
  manifold.euclidean(3),
  fn = function(beta, data) {
    (data$X$matmul(beta) - data$y)$square()$mean() / 2
  },
  equality = function(beta, data) (beta[2:3]$sum() - 3)$reshape(1),
  inequality = function(beta, data) -beta[2:3],
  data = registered
)
```

[`riem.problem.constrained()`](../reference/riem.problem.constrained.html)
uses automatic differentiation for the objective gradient and constraint
adjoints here. The augmented-Lagrangian solver repeatedly solves smooth inner
problems and updates its multipliers.

## Solve and inspect termination

```{r solve}
initial <- torch_tensor(c(0, 1.5, 1.5), dtype = torch_float64(), device = "cpu")
fit <- riem.optimize(
  problem, initial, method = "augmented_lagrangian",
  control = list(max_iterations = 30, gradient_tolerance = 1e-7,
                 feasibility_tolerance = 1e-7,
                 inner_control = list(max_iterations = 100)),
  device = "cpu"
)
centered_coefficients <- as.numeric(fit$point)
coefficients <- c(
  intercept = centered_coefficients[1] + response_center -
    sum(centered_coefficients[2:3] * predictor_center),
  log_diameter = centered_coefficients[2],
  log_height = centered_coefficients[3]
)
round(coefficients, 6)
data.frame(termination = fit$termination, as.data.frame(fit$kkt))
```

`converged_kkt` means the feasibility, stationarity, complementarity, and dual
feasibility tests passed. An iteration budget or an inner-solver failure is a
different outcome and must not be reported as convergence. These residuals
usually give local optimality conditions; the independent calculation below
also takes advantage of this particular problem's convexity.

## Visualize the fit

```{r fit-plot, echo=FALSE, fig.width=6, fig.height=6, fig.alt="Observed timber volume against exponentiated fitted log volume for 31 trees, with a diagonal perfect-fit line.", fig.cap="The fit is optimized in log-volume units. Exponentiating a fitted log value does not correct for retransformation bias."}
fitted_log_volume <- drop(X %*% centered_coefficients) + response_center
fitted_volume <- exp(fitted_log_volume)
limits <- range(c(trees$Volume, fitted_volume))
local({
  old_par <- par(pty = "s")
  on.exit(par(old_par))
  plot(trees$Volume, fitted_volume, pch = 19, col = "#267C74",
       xlim = limits, ylim = limits, asp = 1,
       xlab = "Observed timber volume (cubic feet)",
       ylab = "Exponentiated fitted log volume (cubic feet)",
       main = "Constrained volume model")
  abline(0, 1, lty = 2, col = "#718494", lwd = 2)
})
```

## Independent verification

Eliminate the height slope by setting \(b_H=3-b_D\). Let
\(z_i=\log D_i-\log H_i\) and \(w_i=\log V_i-3\log H_i\).
After profiling out the intercept, this is simple least squares in \(b_D\),
restricted to \([0,3]\). The unrestricted slope is a ratio of centered cross
products; clipping it to this interval gives the constrained minimum. This
reference uses base R rather than another call to the riemtorch solver.

```{r independent-verification}
z <- log_predictors[, 1] - log_predictors[, 2]
w <- log_volume - 3 * log_predictors[, 2]
z_centered <- z - mean(z)
w_centered <- w - mean(w)
reference_diameter <- min(3, max(0, sum(z_centered * w_centered) /
                                  sum(z_centered^2)))
reference <- c(intercept = mean(w) - reference_diameter * mean(z),
               log_diameter = reference_diameter,
               log_height = 3 - reference_diameter)
coefficient_error <- max(abs(coefficients - reference))

# Independently evaluate the Lagrangian gradient and all constraint residuals.
beta <- centered_coefficients
lambda <- as.numeric(fit$multipliers$equality)
mu <- as.numeric(fit$multipliers$inequality)
objective_gradient <- drop(crossprod(X, drop(X %*% beta) - y)) / nrow(X)
lagrangian_gradient <- objective_gradient + c(0, lambda, lambda) - c(0, mu)
independent_kkt <- c(
  feasibility = max(abs(sum(beta[2:3]) - 3), pmax(0, -beta[2:3])),
  stationarity = sqrt(sum(lagrangian_gradient^2)),
  complementarity = max(abs(mu * beta[2:3])),
  dual_feasibility = max(0, -mu)
)
rbind(riemtorch = coefficients, independent = reference)
data.frame(maximum_coefficient_error = coefficient_error,
           as.list(independent_kkt), check.names = FALSE)
stopifnot(fit$termination == "converged_kkt",
          coefficient_error < 2e-5,
          max(independent_kkt) < 1e-6,
          max(abs(unlist(fit$kkt) - independent_kkt)) < 1e-8)
```

The fitted slopes are close to the familiar diameter-squared-times-height
relationship. That agreement does not establish that the constraints are
appropriate for another species or population. Predictions
shown here are in-sample and are not a forecast accuracy assessment.

## Choose a device for your own run

All code above uses CPU float64 explicitly so the published result is easy to
reproduce. [`riem.optimize()`](../reference/riem.optimize.html) accepts
`device = "auto"` to discover a suitable device, `device = "cpu"` to override
an available GPU, or an indexed CUDA device on a compatible workstation.
Registered tensors travel with the problem.

```{r device-options, eval=FALSE}
# Change only the device argument in the solve above:
# device = "auto"    # automatic selection while preserving float64
# device = "cpu"     # explicit CPU override
# device = "cuda:0"  # explicit first CUDA GPU; errors if unavailable
```

```{r reproducibility-result, include=FALSE}
example_result <- c(coefficients, objective = fit$objective,
                    coefficient_error = coefficient_error, independent_kkt)
stopifnot(is.numeric(example_result), all(is.finite(example_result)))
```
