---
title: "Summarize MEG covariance features geometrically"
description: "Optimize a log-Euclidean covariance mean for 216 transformed MEG epochs from Riemann's ERP data."
output: rmarkdown::html_document
resource_files:
  - data/riemann-ERP.rds
---

```{r setup, include=FALSE}
library(riemtorch)
library(torch)
if (!torch_is_installed()) {
  stop("This worked example requires the torch runtime. Run torch::install_torch() first.")
}
torch_set_num_threads(1)
torch_manual_seed(7103)
set.seed(7103)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8,
                      fig.height = 4, fig.align = "center", dpi = 120)
```

How can we summarize a collection of covariance matrices without leaving the
positive-definite cone? This example uses 216 matrices from `Riemann::ERP` to
optimize a log-Euclidean mean, compare it with the arithmetic mean, and describe
how far individual epochs lie from their common geometric center.

Start with [the two-matrix SPD introduction](example-spd-mean.html) for the basic
interface. [Average covariance matrices geometrically](example-covariance-means.html)
compares log-Euclidean and affine-invariant means on smaller matrices. Here we
focus on a larger dataset and an exact reformulation that makes optimization
inexpensive. The [geometry guide](geometry-and-metrics.html) explains the metric
implemented by `manifold.spd()`.

## Data: transformed epochs from one participant

The website bundles a snapshot of `ERP` from Riemann 0.1.7 at commit
`e847692576f8b250c8140e1343b145f72caa379f`. Its 216 positive-definite, 32-by-32
matrices describe repeated epochs from **one participant** in the MNE sample
audiovisual experiment. They are not observations from 216 participants.

The coordinates are **16 class-prototype features and 16 filtered trial
features**, produced by xDAWN spatial filtering and prototype augmentation.
They are not 32 original EEG or MEG sensors. The historical example filename
says EEG, but its processing selects MEG and excludes EEG. The representation
was learned using labels from a separate 72-epoch training set: it is not an
unsupervised transformation of the raw recordings.

The stored labels are LA, LV, RA, and RV (left/right auditory/visual events).
We use them only to color a descriptive plot, not to train or evaluate a
classifier. The [data-source page](example-data-sources.html) records the
snapshot, attribution, corrected provenance, and CC0 source-data notice.
Neither Riemann nor a dataset download is needed when building this article.

```{r data}
erp <- readRDS("data/riemann-ERP.rds")
covariances <- erp$covariance
labels <- erp$label
p <- dim(covariances)[1L]
n <- dim(covariances)[3L]
stopifnot(identical(dim(covariances), c(32L, 32L, 216L)),
          length(labels) == n, !anyNA(labels), all(is.finite(covariances)))
table(labels)
```

The data retain very small numerical values from the original feature
construction. Small absolute eigenvalues alone do not indicate poor
conditioning. Divide **every matrix by the same positive scalar**: the mean
diagonal value across all matrices. This improves the numerical scale without
altering the log-Euclidean distances, because
\(\log(C/s)=\log C-(\log s)I\). It also preserves affine-invariant distances.
Multiplying an estimated mean by `common_scale` restores the original scale.
We add no diagonal jitter, shrinkage, or per-epoch normalization.

```{r scaling}
common_scale <- mean(vapply(seq_len(n), function(i) {
  mean(diag(covariances[, , i]))
}, numeric(1)))
stopifnot(is.finite(common_scale), common_scale > 0)
scaled_covariances <- covariances / common_scale
eigenvalues <- vapply(seq_len(n), function(i) {
  eigen(scaled_covariances[, , i], symmetric = TRUE, only.values = TRUE)$values
}, numeric(p))
stopifnot(min(eigenvalues) > 0)
data.frame(common_scale = common_scale,
           smallest_scaled_eigenvalue = min(eigenvalues),
           largest_condition_number = max(eigenvalues[1, ] / eigenvalues[p, ]))
```

## Objective and geometry

For \(C_i\in\mathrm{SPD}(32)\), minimize

\[
 F(G)=\frac{1}{2n}\sum_{i=1}^n
      \|\log G-\log C_i\|_F^2.
\]

Let \(L_i=\log C_i\), \(\bar L=n^{-1}\sum_i L_i\), and
\(v=n^{-1}\sum_i\|L_i-\bar L\|_F^2\). The **same objective**, including its
constant term, is

\[
 F(G)=\tfrac12\|\log G-\bar L\|_F^2+\tfrac12v.
\]

Compute the matrix logarithms once using `riem.matrix.function()`. The optimizer
then needs one mean logarithm and one scalar, rather than all 216 matrices on
every iteration. This is an exact sufficient-statistic reduction, not a
minibatch approximation. The sample axis is separate from the two matrix axes.

```{r sufficient-statistics}
M <- manifold.spd(p, metric = "lerm")
covariance_tensor <- torch_tensor(aperm(scaled_covariances, c(3, 1, 2)),
                                  dtype = torch_float64(), device = "cpu")
log_tensor <- riem.matrix.function(covariance_tensor, "log")
mean_log <- log_tensor$mean(dim = 1)
dispersion <- ((log_tensor - mean_log$unsqueeze(1))^2)$sum() / n
problem <- riem.problem(
  M,
  fn = function(G, data) {
    difference <- riem.matrix.function(G, "log") - data$mean_log
    ((difference^2)$sum() + data$dispersion) / 2
  },
  rgrad = function(G, data) {
    log_G <- riem.matrix.function(G, "log")
    riem.matrix.frechet(log_G, log_G - data$mean_log, "exp")
  },
  data = list(mean_log = mean_log, dispersion = dispersion)
)
```

The log-Euclidean metric is Euclidean in logarithmic coordinates. The gradient
there is \(\log G-\bar L\); applying the differential of the matrix exponential
converts it to a tangent vector at \(G\). Both objective tensors are registered
through `data`, so device selection can move them with the problem.

## Solve from the arithmetic mean

The arithmetic mean is positive definite and provides a familiar initial point.
We let `riem.optimize()` find the geometric mean rather than supplying its
closed-form solution as the initialization.

```{r solve}
arithmetic <- apply(scaled_covariances, c(1, 2), mean)
initial <- torch_tensor(arithmetic, dtype = torch_float64(), device = "cpu")
fit <- riem.optimize(problem, initial, method = "steepest_descent", device = "cpu",
                     control = list(max_iterations = 20,
                                    gradient_tolerance = 1e-7))
geometric <- as.matrix(fit$point)
data.frame(objective = fit$objective, iterations = fit$iterations,
           gradient_norm = fit$gradient_norm, termination = fit$termination)
```

A unit step in logarithmic coordinates solves this quadratic problem; fast
convergence here reflects its special geometry. Read the termination reason
alongside the gradient norm: exhausting an iteration budget would not establish
convergence on another objective.

## Plot the mean and distances

Both heatmaps use the same scale. Feature numbers 1--16 denote prototype
features and 17--32 denote filtered trial features; the dividing lines separate
these groups. The entries are covariances divided by the common scale, not
correlations or original sensor covariances.

```{r erp-mean-heatmaps, fig.cap="Arithmetic and log-Euclidean means of all 216 covariance matrices, divided by a common positive scale. Lines separate prototype and trial feature groups; the color bar gives scaled covariance values.", fig.alt="Two 32-by-32 heatmaps with a shared color bar compare arithmetic and log-Euclidean covariance means. Dotted lines divide the first 16 prototype features from the last 16 trial features.", fig.width=9, fig.height=4}
panels <- list(Arithmetic = arithmetic, `Log-Euclidean` = geometric)
color_limit <- max(abs(unlist(panels)))
old_par <- par(no.readonly = TRUE)
layout(matrix(1:3, nrow = 1), widths = c(1, 1, 0.18))
par(mar = c(4, 4, 3, 1))
for (name in names(panels)) {
  image(seq_len(p), seq_len(p), panels[[name]],
        col = hcl.colors(80, "Blue-Red 3"),
        zlim = c(-color_limit, color_limit), asp = 1,
        xlab = "Feature", ylab = "Feature", main = name)
  abline(v = 16.5, h = 16.5, lty = 3, col = "#333333")
}
par(mar = c(4, 0, 3, 3))
color_values <- seq(-color_limit, color_limit, length.out = 80)
image(1, color_values, matrix(color_values, nrow = 1),
      col = hcl.colors(80, "Blue-Red 3"),
      axes = FALSE, xlab = "", ylab = "")
axis(4, las = 1, cex.axis = 0.8)
box()
par(old_par)
```

The distance to the geometric mean is the Frobenius norm between logarithms.
Compute all distances from the already available logarithms. Stored epoch order
is an index, not elapsed recording time; point colors identify the event labels.

```{r distances}
log_geometric <- riem.matrix.function(fit$point, "log")
distances <- as.numeric(((log_tensor - log_geometric$unsqueeze(1))^2)$
                         sum(dim = c(2, 3))$sqrt())
aggregate(distances, list(label = labels), function(x) {
  c(count = length(x), median = median(x), maximum = max(x))
})
```

```{r erp-epoch-distances, fig.cap="Distances from each transformed epoch to the common log-Euclidean mean; colors show the four stored event labels. This is a descriptive summary of one participant.", fig.alt="A scatterplot shows the log-Euclidean distance of each of 216 MEG epochs from their common covariance mean, colored by LA, LV, RA, and RV event labels.", fig.width=8, fig.height=4}
colors <- c(LA = "#21618C", LV = "#C0652A", RA = "#3B7D58", RV = "#8856A7")
plot(seq_len(n), distances, pch = 19, cex = 0.7,
     col = colors[as.character(labels)],
     ylim = c(min(distances) - 0.1, max(distances) + 0.5),
     xlab = "Stored epoch index", ylab = "Log-Euclidean distance to mean")
legend("topright", legend = names(colors), col = colors, pch = 19,
       ncol = 4, bty = "n", cex = 0.8)
```

## Independent numerical checks

Recalculate all matrix logarithms and the exact mean using base R's symmetric
eigendecomposition. This is independent of the torch matrix kernels used in the
solve. Check the mean, the original full objective, positive definiteness, and
stationarity in logarithmic coordinates.

```{r independent-reference}
matrix_function <- function(C, fun) {
  eig <- eigen((C + t(C)) / 2, symmetric = TRUE)
  eig$vectors %*% diag(fun(eig$values)) %*% t(eig$vectors)
}
reference_logs <- lapply(seq_len(n), function(i) {
  matrix_function(scaled_covariances[, , i], log)
})
reference_mean_log <- Reduce(`+`, reference_logs) / n
reference_mean <- matrix_function(reference_mean_log, exp)
relative_error <- norm(geometric - reference_mean, "F") / norm(reference_mean, "F")
fitted_log <- matrix_function(geometric, log)
stationarity <- norm(fitted_log - reference_mean_log, "F")
full_objective <- mean(vapply(reference_logs, function(L) {
  sum((fitted_log - L)^2) / 2
}, numeric(1)))
objective_error <- abs(full_objective - fit$objective)
minimum_eigenvalue <- min(eigen(geometric, symmetric = TRUE)$values)
stopifnot(relative_error < 1e-7, stationarity < 1e-7,
          objective_error < 1e-8, minimum_eigenvalue > 0,
          riem.belongs(M, fit$point), fit$termination == "converged_gradient")
data.frame(check = c("Relative error against base-R mean",
                     "Independent stationarity norm",
                     "Full/reduced objective difference",
                     "Smallest eigenvalue of fitted mean"),
           value = c(relative_error, stationarity, objective_error, minimum_eigenvalue))
```

As a separate preprocessing check, recalculate the distance between the first
two matrices on their original scale. A common scalar preserves the distance;
changing each epoch's scale separately would define a different analysis.

```{r scale-check}
original_distance <- norm(matrix_function(covariances[, , 1], log) -
                            matrix_function(covariances[, , 2], log), "F")
scaled_distance <- norm(reference_logs[[1]] - reference_logs[[2]], "F")
scale_distance_error <- abs(original_distance - scaled_distance)
stopifnot(scale_distance_error < 1e-8)
scale_distance_error
```

## Interpretation and device choices

The geometric mean minimizes squared log-Euclidean distances; the arithmetic
mean minimizes squared Frobenius distances between the matrices themselves.
Their visual differences reflect this choice of objective. Neither is a
generally superior summary, and the plotted distances are not a statistical
test or evidence of predictive performance. The epochs share a participant and
a supervised feature construction. A new split of these packaged matrices
would not recreate raw-data preprocessing or establish generalization to new
participants.

CPU float64 and one torch thread make this article reproducible. To request
automatic device discovery or override it, change the `device` argument. The
following alternatives are shown without requiring accelerator hardware during
the website build. See [Applications and devices](applications-and-devices.html).

```{r device-options, eval=FALSE}
riem.optimize(problem, initial, device = "auto")
riem.optimize(problem, initial, device = "cpu")
riem.optimize(problem, initial, device = "cuda:0")
```

```{r reproducibility-result, include=FALSE}
example_result <- c(relative_error = relative_error, stationarity = stationarity,
                    objective = fit$objective, objective_error = objective_error,
                    smallest_eigenvalue = minimum_eigenvalue,
                    scale_distance_error = scale_distance_error,
                    common_scale = common_scale,
                    mean = as.vector(geometric), distances = distances)
names(example_result) <- make.unique(names(example_result))
stopifnot(all(is.finite(example_result)))
```
