---
title: "Get started with riemtorch"
description: "Define a constrained tensor objective, solve it, check the result, and choose CPU or GPU execution."
output: rmarkdown::html_document
---

```{r setup, include=FALSE}
library(torch)
library(riemtorch)
runtime_available <- torch::torch_is_installed()
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      eval = runtime_available, fig.width = 7, fig.height = 4)
if (runtime_available) {
  torch_set_num_threads(1)
  torch_manual_seed(7)
}
```

`riemtorch` minimizes objectives that you write with R torch, while keeping the
unknowns on a chosen manifold. A typical workflow has three ingredients:
the geometry, a scalar loss function, and a feasible starting point. The package
obtains derivatives with torch and uses the geometry to turn them into
Riemannian optimization steps.

This tutorial finds the unit vector most aligned with a given direction. The
answer is known, so we can check the objective, the constraint, and the final
point independently.

## Installation

Install the R torch dependency and its numerical runtime once, then install
riemtorch from GitHub. These commands are setup instructions; the documentation
build does not run them.

```{r installation, eval=FALSE}
install.packages(c("remotes", "torch"))
if (!torch::torch_is_installed()) torch::install_torch()
remotes::install_github("kisungyou/riemtorch")
```

To install a local checkout instead, open `riemtorch.Rproj` and use
**Build → Install Package** in RStudio. If riemtorch is already installed,
start with the next section. Reinstall the
package after changing its source code to use those changes in a fresh R session.

```{r runtime-note, echo=FALSE, results='asis', eval=TRUE}
if (!runtime_available) {
  cat("**Examples were not executed:** the torch runtime is unavailable. ",
      "Run the setup above and rebuild this site to see the numerical output.\n")
}
```

## Define the geometry and objective

Let \(a = (1, 2, 3)\). We solve

\[
\min_{x \in \mathbb{R}^3} -a^\mathsf{T}x
\qquad\text{subject to}\qquad \|x\|_2 = 1.
\]

The feasible set is the two-dimensional sphere embedded in three-dimensional
space, specified by `manifold.sphere(3)`. The objective must return one real
scalar torch tensor, with the same precision and device as the point.

```{r define-problem}
library(torch)
library(riemtorch)

sphere <- manifold.sphere(3)
a <- torch_tensor(c(1, 2, 3), dtype = torch_float64())
problem <- riem.problem(
  sphere,
  fn = function(x, data) -torch_sum(x * data),
  data = a,
  label = "Find an aligned unit vector",
  required_operations = "basic"
)

x0 <- torch_tensor(c(0, 1, 0), dtype = torch_float64())
riem.belongs(sphere, x0)
```

The `data` argument registers `a` with the problem. The execution driver can then
move it together with the initial point when a run uses a different device.
Avoid capturing device-bound tensors invisibly inside the loss function.
Here the loss needs only basic tensor operations, so that is the compatibility
requirement we declare. Leave `required_operations` unset for a conservative
probe when the needs of a custom loss are unknown.

## Solve and inspect the result

For reproducible output, this article explicitly runs on the CPU in double
precision. Automatic device selection is explained below.

```{r solve}
fit <- riem.optimize(
  problem, x0,
  method = "conjugate_gradient",
  control = list(max_iterations = 100, gradient_tolerance = 1e-6),
  device = "cpu"
)
fit
```

The returned point is available in `fit$point`. Check both the termination
reason and the constraint: reaching an iteration budget does not mean the
gradient tolerance was met. For this example the exact minimizer is
\(a / \|a\|_2\), and the minimum is \(-\|a\|_2\).

```{r verify-solution}
exact <- a / a$norm()
data.frame(
  termination = fit$termination,
  on_sphere = riem.belongs(sphere, fit$point),
  gradient_norm = fit$gradient_norm,
  objective_gap = fit$objective + a$norm()$item(),
  point_error = (fit$point - exact)$norm()$item()
)
as.numeric(fit$point)
fit$evaluations
```

`converged_gradient` means the metric gradient norm met the requested tolerance.
`max_iterations`, `stopped_small_step`, and `line_search_failed` describe
different stopping conditions and should be interpreted alongside the objective
and residuals. The evaluation counts help distinguish iteration counts from
actual objective and derivative work.

## Compare a second solver

The problem and initial point can be reused unchanged. Here we compare conjugate
gradient with steepest descent on exactly the same small objective.

```{r compare-solvers}
steepest <- riem.optimize(
  problem, x0,
  method = "steepest_descent",
  control = list(max_iterations = 100, gradient_tolerance = 1e-6),
  device = "cpu"
)

fits <- list("Conjugate gradient" = fit, "Steepest descent" = steepest)
data.frame(
  solver = names(fits),
  iterations = vapply(fits, function(z) z$iterations, numeric(1)),
  objective = vapply(fits, function(z) z$objective, numeric(1)),
  gradient_norm = vapply(fits, function(z) z$gradient_norm, numeric(1)),
  termination = vapply(fits, function(z) z$termination, character(1)),
  row.names = NULL
)
```

```{r convergence-plot, fig.alt='Metric gradient norm against iteration for conjugate gradient and steepest descent on a sphere.', fig.cap='Both methods use the same objective and initial point. This small example illustrates how to read a history; it is not a performance benchmark.'}
history_xy <- function(z) {
  data.frame(
    iteration = vapply(z$history, function(h) h$iteration, numeric(1)),
    gradient = vapply(z$history, function(h) h$gradient_norm, numeric(1))
  )
}
paths <- lapply(fits, history_xy)
all_gradients <- unlist(lapply(paths, function(z) pmax(z$gradient, 1e-12)))
plot(paths[[1]]$iteration, pmax(paths[[1]]$gradient, 1e-12),
     type = "o", pch = 16, col = "#147D92", log = "y",
     xlim = range(unlist(lapply(paths, function(z) z$iteration))),
     ylim = range(all_gradients), xlab = "Iteration",
     ylab = "Metric gradient norm")
lines(paths[[2]]$iteration, pmax(paths[[2]]$gradient, 1e-12),
      type = "o", pch = 1, col = "#C15F2D")
abline(h = 1e-6, lty = 3, col = "grey50")
legend("topright", legend = names(fits), col = c("#147D92", "#C15F2D"),
       pch = c(16, 1), lty = 1, bty = "n")
```

## Check derivatives before a larger run

Check the geometry and the loss at the initial point, away from the solution.
The gradient checker compares the directional derivative with finite
differences. It also records a Taylor remainder over several step sizes.

```{r diagnostics}
direction <- torch_tensor(c(1, 0, 0), dtype = torch_float64())
geometry_check <- riem.check.manifold(sphere, x0, direction)
gradient_check <- riem.check.gradient(problem, x0, direction)

geometry_check$checks[, c("check", "error", "status")]
gradient_check$checks[, c("check", "error", "status")]
gradient_check$curves
```

A `pass` is evidence at the tested point and direction, not a proof for the whole
problem. Repeat checks at several points when adding an analytic derivative or
a custom geometry. See [Checking geometry and derivatives](checking-derivatives.html)
for Hessian and adjoint diagnostics.

## Choose CPU or automatic GPU discovery

Managed runs select one compatible device and retain it for the whole run.
With no explicit device argument, the order of precedence is the R option
`riemtorch.device`, the environment variable `RIEMTORCH_DEVICE`, then automatic
discovery. Discovery tries visible CUDA devices, compatible MPS, then CPU,
preserving the requested precision. The default double precision can rule out
an accelerator that only supports single precision for the required operations.

```{r inspect-device}
fit$execution[c("device", "dtype", "reason")]
riem.devices(dtype = "float64", operations = "basic")
```

These alternative calls are shown without execution so that this page builds
on a machine without a GPU:

```{r device-options, eval=FALSE}
# Discover a compatible device, even if an R option or environment default is set.
automatic_fit <- riem.optimize(problem, x0, device = "auto")

# Force CPU execution, including on a workstation with CUDA GPUs.
cpu_fit <- riem.optimize(problem, x0, device = "cpu")

# Use the first visible CUDA GPU; indices start at zero.
gpu_fit <- riem.optimize(problem, x0, device = "cuda:0")

# Set a session default for runs that omit the device argument.
options(riemtorch.device = "cpu")
default_cpu_fit <- riem.optimize(problem, x0)
options(riemtorch.device = NULL)  # Remove the session override.
```

An explicit unavailable or incompatible device raises an error. Selection does
not silently reduce precision or switch devices in the middle of a run.
The [applications and devices guide](applications-and-devices.html) covers
registered tensor data and device-aware training in more detail.

## Where to go next

- [Example gallery](examples.html): ten illustrated introductions and public-data applications.
- [Data sources and reproducibility](example-data-sources.html): local data, provenance, and numerical checks.
- [Geometry and metrics](geometry-and-metrics.html): choose a geometry and its metric.
- [Custom objectives](custom-objectives.html): write losses and register tensor data.
- [Deterministic optimization](deterministic-optimization.html): solver controls and termination.
- [Second-order and least-squares methods](second-order-and-leastsquares.html): Hessian and residual models.
- [Torch training](torch-training.html): optimize manifold-valued module parameters.
